---
title: "Where should robot delivery go next?"
subtitle: "A store-level suitability score built from open data"
author: "Kristian Vepsäläinen"
date: today
lang: en
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: false
execute:
  warning: false
  message: false
  # knitr's chunk cache is deliberately off. The expensive steps — the
  # country extract, the population grid, the postal geometries — are cached
  # on disk by cached() instead, which the analysis controls explicitly.
  # knitr's cache does not track objects across chunks unless every
  # dependency is declared, so editing one chunk silently restored stale
  # objects into the next one.
  cache: false
params:
  radius_m: 3000
  n_shortlist: 20
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(sf)
library(fst)
library(osmextract)
library(geofi)
library(ggplot2)
library(readr)
library(knitr)

sf_use_s2(FALSE)
# cli's dynamic console output is useless in a rendered document and is a
# recurring source of spurious errors during long runs.
options(cli.dynamic = FALSE, cli.ansi = FALSE, cli.unicode = FALSE)

cache_dir <- here("data", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

# Simple on-disk cache: Overpass queries are slow and should not be
# re-run on every render.
cached <- function(name, expr) {
  path <- file.path(cache_dir, paste0(name, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  val <- force(expr)
  # Never cache an empty result. A failed fetch that gets cached looks
  # exactly like a successful fetch of nothing on every later run.
  n <- if (is.null(val)) 0 else NROW(val)
  if (n == 0) {
    warning("Empty result for '", name, "' — not cached.")
    return(val)
  }
  saveRDS(val, path)
  val
}

# Different spatial sources name the geometry column differently ("geom"
# from WFS, "geometry" from most R packages). Normalise it once so that
# downstream code can refer to it by a fixed name.
set_geom_name <- function(x, name = "geometry") {
  current <- attr(x, "sf_column")
  if (identical(current, name)) return(x)
  names(x)[names(x) == current] <- name
  st_geometry(x) <- name
  x
}
```

## The question

In Finland, robot delivery operates out of S Group grocery stores within a
radius of roughly three kilometres, travelling along pedestrian and cycle
paths. The question is which not-yet-served stores should be activated
next, and on what grounds.

The answer is built entirely from open data. That means internal order and
telemetry data is absent, so this is a screening exercise rather than a
final decision. The section [What internal data would change] sets out
which assumptions would be replaced by measurement.

## Data

| Dataset | Source | Used for |
|---|---|---|
| Paavo postal code statistics | Statistics Finland | Population, age structure, income |
| Postal area geometries | Statistics Finland / `geofi` | Allocating population to service areas |
| Stores, pedestrian network, crossings | OpenStreetMap / Overpass | Locations, drivability |
| Robot locations and start dates | Regional co-op press releases | Model validation |

### Paavo

```{r paavo}
paavo_raw <- read_fst(here("data", "paavo", "paavo.fst"))

# Paavo ships Finnish column names containing spaces and a two-letter group
# code in parentheses. Everything downstream uses only the English names on
# the left-hand side of this mapping.
paavo_map <- c(
  zip              = "Postinumeroalue",
  year             = "Vuosi",
  area_m2          = "Postinumeroalueen pinta-ala",
  population       = "Asukkaat yhteensä (HE)",
  mean_age         = "Asukkaiden keski-ikä (HE)",
  age_25_29        = "25-29-vuotiaat (HE)",
  age_30_34        = "30-34-vuotiaat (HE)",
  age_35_39        = "35-39-vuotiaat (HE)",
  age_40_44        = "40-44-vuotiaat (HE)",
  age_45_49        = "45-49-vuotiaat (HE)",
  age_50_54        = "50-54-vuotiaat (HE)",
  age_55_59        = "55-59-vuotiaat (HE)",
  age_60_64        = "60-64-vuotiaat (HE)",
  households       = "Taloudet yhteensä (TE)",
  hh_young_single  = "Nuorten yksinasuvien taloudet (TE)",
  hh_small_child   = "Pienten lasten taloudet (TE)",
  hh_rental        = "Vuokra-asunnoissa asuvat taloudet (TE)",
  hh_pensioner     = "Eläkeläisten taloudet (TE)",
  dwellings        = "Asunnot (RA)",
  dwell_apartment  = "Kerrostaloasunnot (RA)",
  median_income    = "Asukkaiden mediaanitulot (HR)",
  purchasing_power = "Asukkaiden ostovoimakertymä (HR)",
  students         = "Opiskelijat (PT)",
  jobs             = "Työpaikat yhteensä (TP)"
)

missing <- setdiff(paavo_map, names(paavo_raw))
if (length(missing) > 0) {
  stop("Paavo columns missing, fix paavo_map: ",
       paste(missing, collapse = ", "))
}

paavo <- paavo_raw |>
  select(all_of(paavo_map)) |>
  # The file holds several vintages; without this every postal area would
  # enter the join more than once and population would multiply.
  filter(year == max(year, na.rm = TRUE)) |>
  mutate(
    # PxWeb exports often carry the code and the area name in one field
    # ("00100 Helsinki Keskusta"), so the five-digit code is extracted rather
    # than assumed. A silently failed join here empties the whole analysis.
    zip = str_pad(str_extract(as.character(zip), "\\d{5}"), 5, pad = "0"),
    working_age = age_25_29 + age_30_34 + age_35_39 + age_40_44 +
                  age_45_49 + age_50_54 + age_55_59 + age_60_64
  ) |>
  select(-starts_with("age_"))

message("Paavo vintage in use: ", unique(paavo$year),
        " | postal areas: ", nrow(paavo))

# Sanity check on the area unit. Paavo reports area in square metres, so a
# median in the millions is expected; a median in the thousands would mean
# hectares and every density below would be off by four orders of magnitude.
message("Median postal area: ", format(median(paavo$area_m2, na.rm = TRUE),
        big.mark = " "))

zip_geom <- cached("zip_geom", {
  z <- geofi::get_zipcodes(year = 2023)
  # The Statistics Finland WFS calls its geometry column "geom", other
  # sources call it "geometry". The sf geometry column is sticky in
  # select(), so it must never be named there; instead the name is
  # normalised once, here, and the rest of the document can rely on it.
  z <- set_geom_name(z)
  z |>
    select(zip = posti_alue) |>
    st_transform(3067)
})

zips <- zip_geom |>
  left_join(paavo, by = "zip")

# Fail loudly on a bad join. Without this the population column is all NA,
# every downstream filter empties out, and the document renders with correct
# code and blank tables.
matched <- mean(!is.na(zips$population))
if (matched < 0.8) {
  stop(sprintf(
    "Only %.0f%% of postal areas matched Paavo. Check the zip formats: %s vs %s",
    100 * matched,
    paste(head(zip_geom$zip, 3), collapse = ", "),
    paste(head(paavo$zip, 3), collapse = ", ")
  ))
}

zips <- zips |>
  filter(!is.na(population), population > 0) |>
  st_make_valid() |>
  mutate(zip_area = as.numeric(st_area(geometry)))
```

Paavo reports population by postal area, but a service area is a circle
around a store, so the two have to be reconciled. An earlier version of
this analysis allocated population by overlapping area, which assumes
people are spread evenly inside a postal area. That assumption failed
visibly in the north: a 3 km circle around the store in Kittilä was
credited with 72 residents, because the postal area covers thousands of
square kilometres and almost everyone in it lives in the centre.

Absolute population is therefore taken from the 1 km population grid, which
records where people actually live. Paavo is still used, but only for
composition — the working-age share and the share of households that order
groceries — because ratios are far less sensitive to the even-spread
assumption than counts are.

### Stores from OpenStreetMap

```{r stores}
# S Group chains. Sale and S-market have served as robot origin points;
# Prisma is included because its assortment supports online orders.
sgroup_pattern <- "^(S-market|Prisma|Sale|Alepa|Food Market Herkku|Sokos Herkku)\\b"

# Stores come from the Geofabrik country extract rather than the Overpass
# API. Overpass is convenient but rate limited, and a work sample whose
# results depend on how a third party's server happens to behave on a given
# afternoon is not reproducible. The extract is downloaded once and cached
# by osmextract; anyone re-running this gets the same rows.
osm_dir <- here("data", "osm")
dir.create(osm_dir, recursive = TRUE, showWarnings = FALSE)

# Stores are selected by name, not by the shop tag. A large share of Finnish
# grocery stores in OSM carry a name but no shop tag, or are mapped only as
# a building outline, so filtering on shop silently dropped roughly half of
# them — including towns that demonstrably have a store. The chain names are
# distinctive enough to do the work on their own.
sgroup_prefixes <- c("S-market", "Prisma", "Sale", "Alepa",
                     "Food Market Herkku", "Sokos Herkku")

name_sql <- function(layer) {
  conds <- paste0("name LIKE '", sgroup_prefixes, "%'", collapse = " OR ")
  # osm_way_id exists only in the multipolygons layer; asking for it on
  # points makes the query fail outright.
  cols <- if (layer == "multipolygons") {
    "osm_id, osm_way_id, name, shop, brand, operator, geometry"
  } else {
    "osm_id, name, shop, brand, operator, geometry"
  }
  paste0("SELECT ", cols, " FROM ", layer, " WHERE ", conds)
}

read_osm_shops <- function(layer) {
  osmextract::oe_get(
    "Finland",
    layer = layer,
    query = name_sql(layer),
    download_directory = osm_dir,
    quiet = FALSE
  ) |>
    mutate(src = layer)
}

stores_raw <- cached("stores_osm", {
  pts <- read_osm_shops("points") |> filter(!is.na(name))

  polys <- read_osm_shops("multipolygons") |>
    filter(!is.na(name)) |>
    st_make_valid() |>
    (\(x) suppressWarnings(st_centroid(x)))()

  message("Shops in extract: ", nrow(pts), " points, ",
          nrow(polys), " polygons")

  bind_rows(pts, polys)
})

stopifnot(nrow(stores_raw) > 0)

stores <- stores_raw |>
  filter(!is.na(name), str_detect(name, sgroup_pattern)) |>
  mutate(
    chain = str_extract(name, sgroup_pattern),
    # In the multipolygons layer osm_id is populated only for relations; a
    # store mapped as a closed way carries its id in osm_way_id and has
    # osm_id = NA. Deduplicating on osm_id alone therefore collapsed every
    # building-mapped store in the country into a single row, which is why
    # towns with an obvious store appeared to have none.
    uid = paste(
      src,
      coalesce(as.character(osm_id),
               if ("osm_way_id" %in% names(stores_raw))
                 as.character(osm_way_id) else NA_character_),
      sep = ":"
    )
  ) |>
  select(any_of(c("uid", "name", "brand", "operator", "chain"))) |>
  set_geom_name() |>
  st_transform(3067) |>
  distinct(uid, .keep_all = TRUE)

# A single store often appears more than once: as a node and again as the
# building around it, or as several POIs inside one Prisma (entrance, info
# desk). Collapse anything of the same chain within 100 m into one store.
store_clusters <- stores |>
  st_buffer(100) |>
  st_union() |>
  st_cast("POLYGON") |>
  st_as_sf() |>
  mutate(cluster_id = row_number()) |>
  set_geom_name()

stores <- stores |>
  st_join(store_clusters, join = st_within) |>
  group_by(cluster_id, chain) |>
  slice(1) |>
  ungroup()

message("Stores after deduplication: ", nrow(stores))

# City comes from a spatial join, not from OSM's addr:city tag. That tag is
# absent on a large share of Finnish shop nodes, and because the served /
# not-served flag is keyed on city, a sparse tag would quietly corrupt the
# validation rather than raise an error.
municipalities <- cached("municipalities", {
  geofi::get_municipalities(year = 2023) |> set_geom_name()
})

city_col <- intersect(c("municipality_name_fi", "name_fi", "nimi", "name"),
                      names(municipalities))[1]

muni_geom <- municipalities |>
  select(city = all_of(city_col)) |>
  st_transform(3067)

stores <- stores |>
  st_join(muni_geom, join = st_within)

# Coastline simplification leaves a handful of stores just outside every
# municipality polygon. Left as NA they would silently count as not served,
# so they are assigned to the nearest municipality instead.
if (any(is.na(stores$city))) {
  gap <- is.na(stores$city)
  nearest <- st_nearest_feature(stores[gap, ], muni_geom)
  stores$city[gap] <- muni_geom$city[nearest]
}

message("S Group stores identified: ", nrow(stores),
        " | without a city match: ", sum(is.na(stores$city)))

stores |>
  st_drop_geometry() |>
  count(chain, sort = TRUE) |>
  kable(caption = "Identified stores by chain")
```

Chain identification relies on OSM `brand` and `name` tags, whose coverage
varies. If a chain looks clearly undercounted above, that is a tagging
problem rather than a real absence, and it should be fixed before the
results are read.

### Current robot locations

```{r served}
# Compiled from regional co-op and SOK press releases: 47 municipalities,
# each with the date of its first robot delivery and the source it came
# from. This list is the model's only validation anchor, so it is kept as a
# data file with citations rather than buried in code.
served_towns <- readr::read_csv(
  here("data", "served", "s-ryhma_ruokarobotti_kunnat_2026.csv"),
  show_col_types = FALSE
)

# Start dates arrive at three precisions: full date, year and month, or year
# only. The coarser ones are placed mid-period, which is accurate enough for
# a rank correlation and wrong for anything finer.
parse_start <- function(x) {
  dashes <- str_count(x, "-")
  as.Date(case_when(
    dashes == 2 ~ x,
    dashes == 1 ~ paste0(x, "-15"),
    TRUE        ~ paste0(x, "-07-01")
  ))
}

served_towns <- served_towns |>
  mutate(
    start = parse_start(aloituspaiva),
    # Rows whose source could not be pinned to a specific press release are
    # marked in the source field. They are kept, because leaving a genuinely
    # served town out of the anchor is the worse error, but they are
    # separable so their influence can be measured instead of assumed.
    confirmed = !str_detect(lahde_id, regex("vahvistamaton", ignore_case = TRUE)),
    city = kunta
  )

stores <- stores |>
  mutate(served = city %in% served_towns$city)

message("Served municipalities on file: ", nrow(served_towns),
        " | stores in them: ", sum(stores$served))
```

## Scoring

Five components, each normalised to the unit interval. The weights are
explicit and live in one place, so that anyone can dispute them
individually. Demand and customer mix are kept apart deliberately: a large
suburb of homeowning pensioners and a small student district can produce
the same headcount and very different order volumes.

```{r components}
weights <- c(demand       = 0.30,  # how many people are within reach
             customer_mix = 0.15,  # how many of them are likely orderers
             drivability  = 0.25,  # can the robot actually get there
             density      = 0.20,  # deliveries per kilometre travelled
             crossings    = 0.10)  # road crossings as a risk discount

buffers <- stores |>
  st_buffer(params$radius_m) |>
  mutate(buffer_id = row_number())

# Absolute population comes from the 1 km population grid, not from postal
# areas. Area weighting assumes people are spread evenly inside a postal
# area; in northern municipalities one postal area can cover thousands of
# square kilometres with everyone living in the centre, so a 3 km circle was
# credited with a few dozen residents where the real figure is thousands.
# The grid measures where people actually are.
pop_grid <- cached("pop_grid", {
  geofi::get_population_grid(year = 2022, resolution = 1) |>
    set_geom_name() |>
    st_transform(3067)
})

grid_pop_col <- intersect(
  c("vaesto", "population", "he_vakiy", "vaesto_lkm", "asukkaita"),
  names(pop_grid)
)[1]
if (is.na(grid_pop_col)) {
  stop("No population column found in the grid: ",
       paste(names(pop_grid), collapse = ", "))
}

grid_population <- pop_grid |>
  transmute(grid_pop = .data[[grid_pop_col]]) |>
  # Cells with very few residents are suppressed and can appear as negative
  # sentinel values; they are floored at zero rather than dropped.
  mutate(grid_pop = pmax(coalesce(as.numeric(grid_pop), 0), 0)) |>
  st_centroid() |>
  st_join(buffers |> select(buffer_id), join = st_within) |>
  st_drop_geometry() |>
  filter(!is.na(buffer_id)) |>
  group_by(buffer_id) |>
  summarise(population = sum(grid_pop, na.rm = TRUE), .groups = "drop")

# Paavo is still used, but only for composition. Shares are far less
# sensitive to the even-spread assumption than absolute counts are.
paavo_shares <- st_intersection(
  buffers |> select(buffer_id),
  zips |> select(zip, population, working_age, households, hh_young_single,
                 hh_small_child, hh_rental, dwell_apartment, dwellings,
                 students, jobs, purchasing_power, zip_area)
) |>
  mutate(share = as.numeric(st_area(geometry)) / zip_area) |>
  st_drop_geometry() |>
  group_by(buffer_id) |>
  summarise(
    across(
      c(population, working_age, hh_young_single, hh_small_child, students),
      ~ sum(.x * share, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  transmute(
    buffer_id,
    working_age_share = working_age / pmax(population, 1),
    orderer_share = (hh_young_single + hh_small_child + students) /
                    pmax(population, 1)
  )

pop_in_service_area <- grid_population |>
  inner_join(paavo_shares, by = "buffer_id") |>
  mutate(working_age = population * working_age_share)
```

```{r osm_network}
# The pedestrian network comes from the same local extract, so the whole
# analysis depends on one downloaded file and no live service.
#
# Every store is scored. An earlier version kept only the 120 largest
# service areas to limit the number of network requests; with the data held
# locally that limit has no purpose, and it was actively harmful: the
# largest service areas are the big cities, which are already served, so the
# shortlist of candidates was empty by construction.
candidates <- buffers |>
  left_join(pop_in_service_area, by = "buffer_id") |>
  filter(!is.na(population))

message("Scoring ", nrow(candidates), " stores, of which ",
        sum(candidates$served), " already served")

path_types <- c("footway", "path", "pedestrian", "living_street", "cycleway")

paths <- cached("paths_osm", {
  osmextract::oe_get(
    "Finland",
    layer = "lines",
    query = paste0(
      "SELECT osm_id, highway, geometry FROM lines WHERE highway IN (",
      paste0("'", path_types, "'", collapse = ", "), ")"
    ),
    download_directory = osm_dir,
    quiet = FALSE
  ) |>
    st_transform(3067)
})

crossings <- cached("crossings_osm", {
  osmextract::oe_get(
    "Finland",
    layer = "points",
    query = "SELECT osm_id, highway, geometry FROM points WHERE highway = 'crossing'",
    download_directory = osm_dir,
    quiet = FALSE
  ) |>
    st_transform(3067)
})

message("Network: ", nrow(paths), " path segments, ",
        nrow(crossings), " crossings")

# Kilometres of pedestrian path inside each service area
path_lengths <- st_intersection(
  candidates |> select(buffer_id),
  paths |> select(osm_id)
) |>
  mutate(seg_m = as.numeric(st_length(geometry))) |>
  st_drop_geometry() |>
  group_by(buffer_id) |>
  summarise(path_km = sum(seg_m, na.rm = TRUE) / 1000, .groups = "drop")

# Crossings inside each service area
crossing_counts <- st_join(
  crossings |> select(osm_id),
  candidates |> select(buffer_id),
  join = st_within
) |>
  st_drop_geometry() |>
  filter(!is.na(buffer_id)) |>
  count(buffer_id, name = "n_crossings")

network_metrics <- candidates |>
  st_drop_geometry() |>
  select(buffer_id) |>
  left_join(path_lengths, by = "buffer_id") |>
  left_join(crossing_counts, by = "buffer_id") |>
  mutate(
    path_km = coalesce(path_km, 0),
    n_crossings = coalesce(n_crossings, 0L)
  )
```

```{r scoring}
norm01 <- function(x) {
  # Winsorised at the 5th and 95th percentiles: one large city must not
  # flatten the rest of the distribution to zero.
  lo <- quantile(x, 0.05, na.rm = TRUE)
  hi <- quantile(x, 0.95, na.rm = TRUE)
  pmin(pmax((x - lo) / (hi - lo), 0), 1)
}

scored <- candidates |>
  st_drop_geometry() |>
  left_join(network_metrics, by = "buffer_id") |>
  mutate(
    area_km2      = pi * (params$radius_m / 1000)^2,
    pop_density   = population / area_km2,
    path_density  = path_km / area_km2,
    crossing_rate = n_crossings / pmax(path_km, 1),

    c_demand       = norm01(working_age),
    c_customer_mix = norm01(orderer_share),
    c_drivability  = norm01(path_density),
    c_density      = norm01(pop_density),
    # Crossings are a risk, so this component is inverted
    c_crossings    = 1 - norm01(crossing_rate),

    score = weights["demand"]       * c_demand +
            weights["customer_mix"] * c_customer_mix +
            weights["drivability"]  * c_drivability +
            weights["density"]      * c_density +
            weights["crossings"]    * c_crossings
  )
```

## Validation

The only external check available is whether the score recognises
locations already in operation. The model was not built to identify them.

```{r validation}
validation <- scored |>
  mutate(rank_pos = rank(-score))

auc_for <- function(towns) {
  flag <- validation$city %in% towns
  p <- validation$score[flag]; n <- validation$score[!flag]
  if (length(p) == 0 || length(n) == 0) return(NA_real_)
  mean(outer(p, n, ">")) + 0.5 * mean(outer(p, n, "=="))
}

auc <- auc_for(served_towns$city)
auc_confirmed <- auc_for(served_towns$city[served_towns$confirmed])

tibble(
  Anchor = c("All towns on file", "Confirmed sources only"),
  Towns = c(nrow(served_towns), sum(served_towns$confirmed)),
  AUC = round(c(auc, auc_confirmed), 3)
) |>
  kable(caption = "Sensitivity of the validation to unconfirmed rows")

validation |>
  group_by(served) |>
  summarise(
    n = n(),
    median_score = round(median(score), 3),
    median_rank  = round(median(rank_pos)),
    .groups = "drop"
  ) |>
  kable(caption = sprintf("Separation, AUC = %.2f", auc))

# served_towns uses "kunta"; validation uses "city". They are joined by name
# throughout, so a municipality renamed or misspelled in either source drops
# out silently rather than raising an error.
# Served towns with no store in the model drop out of the validation
# entirely. They are listed rather than imputed: placing a notional store at
# a town centre would produce service-area populations and path lengths that
# look measured but are invented, which is a worse failure than a smaller
# validation set.
unmatched <- setdiff(served_towns$city, unique(validation$city))
if (length(unmatched) > 0) {
  message(length(unmatched), " served towns have no S Group store in OSM ",
          "and are excluded from the validation: ",
          paste(unmatched, collapse = ", "))
}

# A stronger test than separation: if the score captures what actually
# drives siting decisions, towns that adopted earlier should score higher.
# This is a rank correlation over the 47 served municipalities, and unlike
# the AUC it cannot be satisfied by simply ranking big cities first.
adoption <- validation |>
  filter(served) |>
  group_by(city) |>
  summarise(town_score = median(score), .groups = "drop") |>
  inner_join(served_towns |> select(city, start), by = "city")

rho <- cor(adoption$town_score, as.numeric(adoption$start),
           method = "spearman", use = "complete.obs")

tibble(
  `Served towns matched` = nrow(adoption),
  `Spearman rho (score vs adoption date)` = round(rho, 3)
) |>
  kable(caption = "Do earlier adopters score higher?")

# A single anchor: according to public reporting, the S-market in central
# Siilinjärvi has been the most active robot origin point in Finland.
validation |>
  filter(city == "Siilinjärvi") |>
  select(name, city, population, path_km, score, rank_pos) |>
  kable(caption = "Where Siilinjärvi lands in the model")
```

If the AUC sits near 0.5, the model does not separate and its components
are the wrong ones. If it is clearly above 0.7, features computed from
open data explain existing siting decisions well enough that the same
features can be used to propose the next ones.

The rank correlation is the harder of the two tests. A negative rho means
higher-scoring towns were served earlier, which is the direction expected
if the score reflects the same considerations the rollout followed. A rho
near zero means the score separates served from unserved but says nothing
about the order in which sites were chosen, which would be worth reporting
as it stands rather than tuning away.

One caveat applies to both tests: the list records each municipality's
first robot delivery, not every store subsequently activated. A store in a
served municipality is therefore treated as served even if robots do not
run from it, which makes both tests conservative.

## Leading candidates

```{r shortlist}
shortlist <- scored |>
  filter(!served) |>
  arrange(desc(score)) |>
  slice_head(n = params$n_shortlist) |>
  transmute(
    Store = name, City = city,
    `Population 3 km` = round(population),
    `Path km` = round(path_km),
    Score = round(score, 3)
  )

kable(shortlist, caption = "Best not-yet-served stores")
```

```{r rollup}
# The commercial decision is made with a regional co-op, not one store at
# a time.
scored |>
  filter(!served) |>
  group_by(City = city) |>
  summarise(
    Stores = n(),
    `Combined population` = round(sum(population)),
    `Median score` = round(median(score), 3),
    .groups = "drop"
  ) |>
  arrange(desc(`Median score`)) |>
  slice_head(n = 15) |>
  kable(caption = "By town: how many stores come with one negotiation")
```

```{r map}
#| fig-height: 7
ggplot() +
  geom_sf(data = municipalities, fill = "grey96", colour = "white",
          linewidth = 0.1) +
  geom_sf(data = scored |>
            left_join(select(buffers, buffer_id), by = "buffer_id") |>
            st_as_sf() |> st_centroid(),
          aes(colour = score, shape = served), size = 2) +
  scale_colour_viridis_c(option = "magma", direction = -1) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4),
                     labels = c("Candidate", "In operation")) +
  labs(colour = "Score", shape = NULL) +
  theme_void()
```

## What internal data would change

Three places where open data stands in for measurement, and where the
score would likely shift materially:

**Actual delivery time would replace path density.** Kilometres of
pedestrian path is a crude proxy for what is really wanted: whether the
robot arrives within a time that keeps the customer happy. Realised
travel time from telemetry, modelled as a function of network structure,
would calibrate this component directly.

**Order volume would replace population count.** Working-age population
within the radius is a proxy for demand. Order density per resident in
existing service areas would show which demographic features actually
predict demand, and the demand component could be estimated rather than
weighted by hand.

**Cost per delivery would settle the weights.** The weights here are set
by judgement. With unit economics data they could be chosen so that the
score predicts contribution rather than volume, which would most likely
move the shortlist toward dense smaller towns.

Winter is absent from this version entirely. Days of snow cover and days
below freezing, from the Finnish Meteorological Institute's open data,
would be a natural fifth component, and it is the one factor no
international benchmark accounts for.

## Limitations

- Population comes from the 1 km grid, whose cells are suppressed where
  very few people live; those cells are counted as zero, which understates
  the most sparsely settled service areas slightly.
- OSM coverage varies regionally; pedestrian network digitisation is
  better in cities than in smaller towns, which biases the score toward
  cities systematically.
- Gradient and surface quality are absent. They determine whether the
  robot can physically get there and would require an elevation model.
- The list of current locations is compiled from press releases and may
  be incomplete, which weakens the validation in both directions.
- Stores are found by chain name in OpenStreetMap. Where a store is absent
  or unnamed in OSM, its town cannot enter the validation at all; those
  towns are listed in the output rather than filled in with a notional
  location.
- Store picking capacity is a binding constraint that open data cannot
  see at all.
