---
title: "Where should robot delivery go next?"
subtitle: "A store-level suitability score built from open data"
author: "Kristian Vepsäläinen"
date: today
lang: en
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: false
execute:
  warning: false
  message: false
  cache: true
params:
  radius_m: 3000
  n_shortlist: 20
  overpass_timeout: 300
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(sf)
library(fst)
library(osmdata)
library(geofi)
library(ggplot2)
library(knitr)

sf_use_s2(FALSE)

cache_dir <- here("data", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

# Simple on-disk cache: Overpass queries are slow and should not be
# re-run on every render.
cached <- function(name, expr) {
  path <- file.path(cache_dir, paste0(name, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  val <- force(expr)
  saveRDS(val, path)
  val
}
```

## The question

In Finland, robot delivery operates out of S Group grocery stores within a
radius of roughly three kilometres, travelling along pedestrian and cycle
paths. The question is which not-yet-served stores should be activated
next, and on what grounds.

The answer is built entirely from open data. That means internal order and
telemetry data is absent, so this is a screening exercise rather than a
final decision. The section [What internal data would change] sets out
which assumptions would be replaced by measurement.

## Data

| Dataset | Source | Used for |
|---|---|---|
| Paavo postal code statistics | Statistics Finland | Population, age structure, income |
| Postal area geometries | Statistics Finland / `geofi` | Allocating population to service areas |
| Stores, pedestrian network, crossings | OpenStreetMap / Overpass | Locations, drivability |
| Current robot locations | Regional co-op press releases | Model validation |

### Paavo

```{r paavo}
paavo_raw <- read_fst(here("data", "paavo", "paavo.fst"))

# Paavo ships Finnish column names containing spaces and a two-letter group
# code in parentheses. Everything downstream uses only the English names on
# the left-hand side of this mapping.
paavo_map <- c(
  zip              = "Postinumeroalue",
  year             = "Vuosi",
  area_m2          = "Postinumeroalueen pinta-ala",
  population       = "Asukkaat yhteensä (HE)",
  mean_age         = "Asukkaiden keski-ikä (HE)",
  age_25_29        = "25-29-vuotiaat (HE)",
  age_30_34        = "30-34-vuotiaat (HE)",
  age_35_39        = "35-39-vuotiaat (HE)",
  age_40_44        = "40-44-vuotiaat (HE)",
  age_45_49        = "45-49-vuotiaat (HE)",
  age_50_54        = "50-54-vuotiaat (HE)",
  age_55_59        = "55-59-vuotiaat (HE)",
  age_60_64        = "60-64-vuotiaat (HE)",
  households       = "Taloudet yhteensä (TE)",
  hh_young_single  = "Nuorten yksinasuvien taloudet (TE)",
  hh_small_child   = "Pienten lasten taloudet (TE)",
  hh_rental        = "Vuokra-asunnoissa asuvat taloudet (TE)",
  hh_pensioner     = "Eläkeläisten taloudet (TE)",
  dwellings        = "Asunnot (RA)",
  dwell_apartment  = "Kerrostaloasunnot (RA)",
  median_income    = "Asukkaiden mediaanitulot (HR)",
  purchasing_power = "Asukkaiden ostovoimakertymä (HR)",
  students         = "Opiskelijat (PT)",
  jobs             = "Työpaikat yhteensä (TP)"
)

missing <- setdiff(paavo_map, names(paavo_raw))
if (length(missing) > 0) {
  stop("Paavo columns missing, fix paavo_map: ",
       paste(missing, collapse = ", "))
}

paavo <- paavo_raw |>
  select(all_of(paavo_map)) |>
  # The file holds several vintages; without this every postal area would
  # enter the join more than once and population would multiply.
  filter(year == max(year, na.rm = TRUE)) |>
  mutate(
    zip = str_pad(as.character(zip), 5, pad = "0"),
    working_age = age_25_29 + age_30_34 + age_35_39 + age_40_44 +
                  age_45_49 + age_50_54 + age_55_59 + age_60_64
  ) |>
  select(-starts_with("age_"))

message("Paavo vintage in use: ", unique(paavo$year),
        " | postal areas: ", nrow(paavo))

# Sanity check on the area unit. Paavo reports area in square metres, so a
# median in the millions is expected; a median in the thousands would mean
# hectares and every density below would be off by four orders of magnitude.
message("Median postal area: ", format(median(paavo$area_m2, na.rm = TRUE),
        big.mark = " "))

zip_geom <- cached("zip_geom", {
  geofi::get_zipcodes(year = 2023) |>
    select(zip = posti_alue, geom = geometry) |>
    st_transform(3067)
})

zips <- zip_geom |>
  left_join(paavo, by = "zip") |>
  filter(!is.na(population), population > 0) |>
  st_make_valid() |>
  mutate(zip_area = as.numeric(st_area(geometry)))
```

Paavo reports population by postal area, but a service area is a circle
around a store. The two are intersected and population is allocated in
proportion to overlapping area. This assumes population is spread evenly
within a postal area, which is wrong in sparsely settled regions and
overstates stores on the edge of large areas. The 250 m population grid
would fix this at the cost of a heavier download and more processing; in
this version the assumption is stated rather than hidden.

### Stores from OpenStreetMap

```{r stores}
# S Group chains. Sale and S-market have served as robot origin points;
# Prisma is included because its assortment supports online orders.
sgroup_brands <- c("S-market", "Prisma", "Sale", "Alepa", "Food Market Herkku")

fetch_stores <- function(brand) {
  opq(bbox = c(19.0, 59.7, 31.6, 70.1), timeout = params$overpass_timeout) |>
    add_osm_features(features = c(
      sprintf('"brand"="%s"', brand),
      sprintf('"name"~"^%s"', brand)
    )) |>
    osmdata_sf()
}

stores_raw <- cached("stores_osm", {
  map(sgroup_brands, function(b) {
    res <- tryCatch(fetch_stores(b), error = function(e) NULL)
    if (is.null(res)) return(NULL)
    pts <- res$osm_points
    polys <- res$osm_polygons
    bind_rows(
      if (!is.null(pts) && nrow(pts) > 0) st_as_sf(pts) else NULL,
      if (!is.null(polys) && nrow(polys) > 0) st_centroid(st_as_sf(polys)) else NULL
    ) |>
      mutate(chain = b)
  }) |>
    bind_rows()
})

stores <- stores_raw |>
  filter(!is.na(name)) |>
  select(any_of(c("osm_id", "name", "brand", "operator", "addr:city", "chain"))) |>
  rename(city = any_of("addr:city")) |>
  st_transform(3067) |>
  distinct(name, chain, .keep_all = TRUE)

# Check this before trusting any number downstream: OSM brand and name
# tagging is uneven, and a misclassified chain propagates straight into
# the result.
stores |>
  st_drop_geometry() |>
  count(chain, sort = TRUE) |>
  kable(caption = "Identified stores by chain")
```

Chain identification relies on OSM `brand` and `name` tags, whose coverage
varies. If a chain looks clearly undercounted above, that is a tagging
problem rather than a real absence, and it should be fixed before the
results are read.

### Current robot locations

```{r served}
# Compiled from regional co-op and SOK press releases. Complete the gaps;
# this list is the model's only validation anchor.
served_towns <- tribble(
  ~city,           ~coop,
  "Helsinki",      "HOK-Elanto",
  "Espoo",         "HOK-Elanto",
  "Vantaa",        "HOK-Elanto",
  "Tampere",       "Pirkanmaa",
  "Turku",         "TOK",
  "Kaarina",       "TOK",
  "Naantali",      "TOK",
  "Raisio",        "TOK",
  "Parainen",      "TOK",
  "Uusikaupunki",  "TOK",
  "Pori",          "Satakunta",
  "Rauma",         "Satakunta",
  "Jyväskylä",     "Keskimaa",
  "Oulu",          "Arina",
  "Lahti",         "Hämeenmaa",
  "Heinola",       "Hämeenmaa",
  "Riihimäki",     "Hämeenmaa",
  "Hämeenlinna",   "Hämeenmaa",
  "Forssa",        "Hämeenmaa",
  "Kotka",         "Kymen Seutu",
  "Kouvola",       "Kymen Seutu",
  "Siilinjärvi",   "PeeÄssä",
  "Kuopio",        "PeeÄssä",
  "Joensuu",       "PKO"
)

stores <- stores |>
  mutate(served = city %in% served_towns$city)
```

## Scoring

Five components, each normalised to the unit interval. The weights are
explicit and live in one place, so that anyone can dispute them
individually. Demand and customer mix are kept apart deliberately: a large
suburb of homeowning pensioners and a small student district can produce
the same headcount and very different order volumes.

```{r components}
weights <- c(demand       = 0.30,  # how many people are within reach
             customer_mix = 0.15,  # how many of them are likely orderers
             drivability  = 0.25,  # can the robot actually get there
             density      = 0.20,  # deliveries per kilometre travelled
             crossings    = 0.10)  # road crossings as a risk discount

buffers <- stores |>
  st_buffer(params$radius_m) |>
  mutate(buffer_id = row_number())

# Population inside the service area, area-weighted
pop_in_service_area <- st_intersection(
  buffers |> select(buffer_id),
  zips |> select(zip, population, working_age, households, hh_young_single,
                 hh_small_child, hh_rental, dwell_apartment, dwellings,
                 students, jobs, purchasing_power, zip_area)
) |>
  mutate(share = as.numeric(st_area(geometry)) / zip_area) |>
  st_drop_geometry() |>
  group_by(buffer_id) |>
  summarise(
    across(
      c(population, working_age, households, hh_young_single, hh_small_child,
        hh_rental, dwell_apartment, dwellings, students, jobs,
        purchasing_power),
      ~ sum(.x * share, na.rm = TRUE)
    ),
    .groups = "drop"
  )
```

```{r osm_network}
# Pedestrian network and crossings are fetched only for leading
# candidates: pulling the whole country from Overpass is neither
# necessary nor polite.
candidates <- buffers |>
  left_join(pop_in_service_area, by = "buffer_id") |>
  filter(!is.na(population)) |>
  arrange(desc(population)) |>
  slice_head(n = 120)

fetch_network <- function(geom_wgs) {
  bb <- st_bbox(geom_wgs)
  q <- opq(bbox = as.numeric(bb), timeout = params$overpass_timeout)
  paths <- tryCatch(
    q |> add_osm_feature("highway",
        c("footway", "path", "pedestrian", "living_street", "cycleway")) |>
      osmdata_sf(),
    error = function(e) NULL)
  crossings <- tryCatch(
    q |> add_osm_feature("highway", "crossing") |> osmdata_sf(),
    error = function(e) NULL)
  list(paths = paths, crossings = crossings)
}

network <- cached("network_osm", {
  candidates |>
    st_transform(4326) |>
    st_geometry() |>
    map(fetch_network)
})

network_metrics <- map2_dfr(network, candidates$buffer_id, function(v, id) {
  path_km <- 0
  if (!is.null(v$paths) && !is.null(v$paths$osm_lines) &&
      nrow(v$paths$osm_lines) > 0) {
    path_km <- v$paths$osm_lines |>
      st_transform(3067) |>
      st_length() |>
      sum() |>
      as.numeric() / 1000
  }
  n_crossings <- 0
  if (!is.null(v$crossings) && !is.null(v$crossings$osm_points)) {
    n_crossings <- nrow(v$crossings$osm_points)
  }
  tibble(buffer_id = id, path_km = path_km, n_crossings = n_crossings)
})
```

```{r scoring}
norm01 <- function(x) {
  # Winsorised at the 5th and 95th percentiles: one large city must not
  # flatten the rest of the distribution to zero.
  lo <- quantile(x, 0.05, na.rm = TRUE)
  hi <- quantile(x, 0.95, na.rm = TRUE)
  pmin(pmax((x - lo) / (hi - lo), 0), 1)
}

scored <- candidates |>
  st_drop_geometry() |>
  left_join(network_metrics, by = "buffer_id") |>
  mutate(
    area_km2      = pi * (params$radius_m / 1000)^2,
    pop_density   = population / area_km2,
    path_density  = path_km / area_km2,
    crossing_rate = n_crossings / pmax(path_km, 1),

    # Share of residents in the household types that dominate quick
    # commerce elsewhere: young people living alone, households with small
    # children, and students.
    orderer_share = (hh_young_single + hh_small_child + students) /
                     pmax(population, 1),

    c_demand       = norm01(working_age),
    c_customer_mix = norm01(orderer_share),
    c_drivability  = norm01(path_density),
    c_density      = norm01(pop_density),
    # Crossings are a risk, so this component is inverted
    c_crossings    = 1 - norm01(crossing_rate),

    score = weights["demand"]       * c_demand +
            weights["customer_mix"] * c_customer_mix +
            weights["drivability"]  * c_drivability +
            weights["density"]      * c_density +
            weights["crossings"]    * c_crossings
  )
```

## Validation

The only external check available is whether the score recognises
locations already in operation. The model was not built to identify them.

```{r validation}
validation <- scored |>
  mutate(rank_pos = rank(-score))

auc <- with(validation, {
  p <- score[served]; n <- score[!served]
  mean(outer(p, n, ">")) + 0.5 * mean(outer(p, n, "=="))
})

validation |>
  group_by(served) |>
  summarise(
    n = n(),
    median_score = round(median(score), 3),
    median_rank  = round(median(rank_pos)),
    .groups = "drop"
  ) |>
  kable(caption = sprintf("Separation, AUC = %.2f", auc))

# A single anchor: according to public reporting, the S-market in central
# Siilinjärvi has been the most active robot origin point in Finland.
validation |>
  filter(str_detect(name, "Siilinjärv")) |>
  select(name, city, population, path_km, score, rank_pos) |>
  kable(caption = "Where Siilinjärvi lands in the model")
```

If the AUC sits near 0.5, the model does not separate and its components
are the wrong ones. If it is clearly above 0.7, features computed from
open data explain existing siting decisions well enough that the same
features can be used to propose the next ones.

## Leading candidates

```{r shortlist}
shortlist <- scored |>
  filter(!served) |>
  arrange(desc(score)) |>
  slice_head(n = params$n_shortlist) |>
  transmute(
    Store = name, City = city,
    `Population 3 km` = round(population),
    `Path km` = round(path_km),
    Score = round(score, 3)
  )

kable(shortlist, caption = "Best not-yet-served stores")
```

```{r rollup}
# The commercial decision is made with a regional co-op, not one store at
# a time.
scored |>
  filter(!served) |>
  group_by(City = city) |>
  summarise(
    Stores = n(),
    `Combined population` = round(sum(population)),
    `Median score` = round(median(score), 3),
    .groups = "drop"
  ) |>
  arrange(desc(`Median score`)) |>
  slice_head(n = 15) |>
  kable(caption = "By town: how many stores come with one negotiation")
```

```{r map}
#| fig-height: 7
municipalities <- cached("municipalities", geofi::get_municipalities(year = 2023))

ggplot() +
  geom_sf(data = municipalities, fill = "grey96", colour = "white",
          linewidth = 0.1) +
  geom_sf(data = scored |>
            left_join(select(buffers, buffer_id), by = "buffer_id") |>
            st_as_sf() |> st_centroid(),
          aes(colour = score, shape = served), size = 2) +
  scale_colour_viridis_c(option = "magma", direction = -1) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4),
                     labels = c("Candidate", "In operation")) +
  labs(colour = "Score", shape = NULL) +
  theme_void()
```

## What internal data would change

Three places where open data stands in for measurement, and where the
score would likely shift materially:

**Actual delivery time would replace path density.** Kilometres of
pedestrian path is a crude proxy for what is really wanted: whether the
robot arrives within a time that keeps the customer happy. Realised
travel time from telemetry, modelled as a function of network structure,
would calibrate this component directly.

**Order volume would replace population count.** Working-age population
within the radius is a proxy for demand. Order density per resident in
existing service areas would show which demographic features actually
predict demand, and the demand component could be estimated rather than
weighted by hand.

**Cost per delivery would settle the weights.** The weights here are set
by judgement. With unit economics data they could be chosen so that the
score predicts contribution rather than volume, which would most likely
move the shortlist toward dense smaller towns.

Winter is absent from this version entirely. Days of snow cover and days
below freezing, from the Finnish Meteorological Institute's open data,
would be a natural fifth component, and it is the one factor no
international benchmark accounts for.

## Limitations

- Population is allocated from postal areas by area weighting, which
  overstates edge stores in large, sparsely settled areas.
- OSM coverage varies regionally; pedestrian network digitisation is
  better in cities than in smaller towns, which biases the score toward
  cities systematically.
- Gradient and surface quality are absent. They determine whether the
  robot can physically get there and would require an elevation model.
- The list of current locations is compiled from press releases and may
  be incomplete, which weakens the validation in both directions.
- Store picking capacity is a binding constraint that open data cannot
  see at all.
