---
title: "Where did the trade go?"
subtitle: "Measuring sanctions circumvention exposure across five Nordic countries"
author: "Kristian Vepsäläinen"
date: today
lang: en
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: false
execute:
  warning: false
  message: false
  cache: false
params:
  baseline_years: 3
  last_year: 2025
  reporters: ["FIN", "SWE", "DNK", "NOR", "ISL"]
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(lubridate)
library(ggplot2)
library(knitr)

options(cli.dynamic = FALSE, cli.ansi = FALSE, cli.unicode = FALSE)

cache_dir <- here("data", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

cached <- function(name, expr) {
  path <- file.path(cache_dir, paste0(name, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  val <- force(expr)
  if (NROW(val) == 0) {
    warning("Empty result for '", name, "' — not cached.")
    return(val)
  }
  saveRDS(val, path)
  val
}
```

## The question

Sanctions on exports to Russia apply to the same goods across the EU and,
through aligned national measures, across Norway and Iceland. The rules are
common; exposure to their circumvention is not. A country's exposure
depends on what it exported to Russia before the restrictions, who its
neighbours are, and how much of that trade re-appeared shortly afterwards
in countries that had never bought it in any quantity.

This report measures that re-appearance in five Nordic countries, using
one metric applied identically to each.

## Method

Circumvention through a third country leaves a specific trace in customs
statistics. Exports of a restricted good to Russia fall to near zero, while
exports of the same good to a small number of transit economies rise far
beyond anything those economies could plausibly consume. The goods are not
consumed there; they are forwarded.

Three things make this measurable rather than suggestive:

**The goods are defined by the legislation, not by me.** The restricted
product codes are enumerated in the annexes of the EU sanctions
regulations, and the headline list is already published at the resolution
trade data uses. Part 1 extracts the codes and the date each one became
restricted; this part counts them. Choosing codes by hand would make any
finding an artefact of that choice.

**The comparison is against the country's own past, code by code.** Each
flow is measured against its own baseline in the years before that
particular code was restricted. A country that always traded heavily with
Central Asia is therefore not flagged for continuing to do so, and a code
restricted in 2025 is not judged against a 2022 boundary.

**A control group absorbs the rest.** Trade to partners with no plausible
onward route to Russia grew or shrank over the same period for ordinary
economic reasons. Differencing against that control removes the part of the
change that is simply the world economy.

## Part 1 — What the legislation defines

```{r codes}
# Two lists, because they are two different animals.
#
# The Common High Priority list is carried in the sanctions regulation
# itself, as Annex XL of Regulation 833/2014 (and Annex XXX of 765/2006 for
# Belarus). It holds 50 items in four tiers and — decisively here — it is
# expressed in six-digit HS codes, the same resolution as international
# trade data. It was drawn up jointly by the EU, US, UK and Japan
# specifically to support anti-circumvention work, so it is the list
# enforcement agencies themselves watch. No truncation is needed.
#
# The vehicle codes used for validation come from the industrial goods
# annex instead, at eight digits, and are truncated to HS6. That truncation
# is confined to the validation and never touches the headline measurement.
chp_codes <- tribble(
  ~hs6,     ~description,                ~tier, ~regulation, ~annex, ~effective,
  "TODO",   "TODO — from Annex XL",      1L,    "833/2014",  "XL",   "TODO",
) |>
  mutate(list = "CHP", cn8_truncated = FALSE)

anchor_codes_tbl <- tribble(
  ~hs6,     ~description,                              ~regulation, ~annex,  ~effective,
  "870421", "Motor vehicles for goods, diesel, <=5t",  "833/2014",  "XXIII", "TODO",
  "870422", "Motor vehicles for goods, diesel, 5-20t", "833/2014",  "XXIII", "TODO",
  "871620", "Trailers and semi-trailers",              "833/2014",  "XXIII", "TODO",
) |>
  mutate(tier = NA_integer_, list = "Industrial", cn8_truncated = TRUE)

restricted_codes <- bind_rows(chp_codes, anchor_codes_tbl) |>
  mutate(effective = as.Date(effective))

# A missing effective date is not a cosmetic gap: without it the code has no
# baseline and would silently be dropped or, worse, compared against the
# wrong years.
stopifnot(!any(is.na(restricted_codes$effective)),
          !any(restricted_codes$hs6 == "TODO"))

restricted_codes |>
  count(list, annex, cn8_truncated, name = "codes") |>
  kable(caption = "Restricted goods carried into the measurement")
```

Codes carry their own start date because the restrictions did not arrive at
once. Twenty-one sanctions packages had been adopted by August 2026, each
one amending the annexes: entries added, codes refined, thresholds revised.
A code restricted from 2023 and a code restricted from 2025 cannot share a
single before-and-after boundary without mixing restricted and unrestricted
years inside the same comparison.

```{r windows}
# Each code gets its own baseline and its own post period. The year the
# restriction takes effect is dropped: it is part restricted and part not,
# and assigning it to either side would bias the comparison in a direction
# that depends on the month of adoption.
code_windows <- restricted_codes |>
  mutate(
    eff_year  = year(effective),
    pre_from  = eff_year - params$baseline_years,
    pre_to    = eff_year - 1,
    post_from = eff_year + 1,
    post_to   = params$last_year,
    pre_width  = pre_to - pre_from + 1,
    post_width = post_to - post_from + 1
  ) |>
  filter(post_width >= 1)

dropped <- nrow(restricted_codes) - nrow(code_windows)
if (dropped > 0) {
  message(dropped, " code(s) restricted too recently to have a post period ",
          "and excluded.")
}

code_windows |>
  count(eff_year, name = "codes") |>
  arrange(eff_year) |>
  kable(caption = "When each batch of codes became restricted")
```

```{r policy_events}
# Part 1's second output: obligations whose effect Part 2 can test.
# Article 12gb of Regulation 833/2014 is the interesting one, because it
# does not restrict a good — it obliges exporters to take active steps
# against circumvention of the CHP items specifically. If that obligation
# bites, its effect should appear in CHP flows and not in the others, which
# is a sharper prediction than "sanctions reduce trade".
policy_events <- tribble(
  ~date,        ~event,                                              ~source,
  "2022-02-24", "Full-scale invasion; sectoral packages begin",       "TODO",
  "2024-12-26", "Art. 12gb anti-circumvention duty applies (Russia)", "TODO",
) |>
  mutate(date = as.Date(date))

policy_events |> kable(caption = "Obligations tested in Part 2")
```

## Part 2 — What the trade data shows

### Partner groups

```{r partners}
# Transit economies are the ones repeatedly identified in EU and national
# enforcement reporting as onward routes to Russia. The list is a stated
# assumption, not a finding, and the results are re-run without each one in
# the sensitivity section.
transit_hubs <- c("ARM", "KAZ", "KGZ", "GEO", "UZB", "TUR", "ARE", "SRB")

# Controls: economies of comparable income and distance with no land or
# practical onward route to Russia. They absorb ordinary trade growth.
control_partners <- c("MAR", "TUN", "JOR", "THA", "MYS", "CHL", "ZAF")

target <- "RUS"
```

### Data

```{r trade_data}
# UN Comtrade is used for all five reporters rather than Eurostat, because
# Norway and Iceland are outside Comext and a cross-country comparison built
# from two different collection systems would confound the measurement with
# the measurement apparatus. Comext at CN8 is the better source for the
# three EU members alone and belongs in the robustness check.
#
# comtradr::set_primary_comtrade_key() needs a free API key.
library(comtradr)

years <- seq(min(code_windows$pre_from), params$last_year)

fetch_flow <- function(reporter, year) {
  cached(sprintf("trade_%s_%d", reporter, year), {
    ct_get_data(
      reporter       = reporter,
      partner        = c(target, transit_hubs, control_partners),
      commodity_code = unique(code_windows$hs6),
      start_date     = year,
      end_date       = year,
      flow_direction = "export",
      freq           = "A"
    )
  })
}

trade <- expand_grid(reporter = params$reporters, year = years) |>
  pmap(function(reporter, year) {
    out <- tryCatch(fetch_flow(reporter, year), error = function(e) {
      warning(reporter, " ", year, ": ", conditionMessage(e))
      NULL
    })
    if (is.null(out) || NROW(out) == 0) return(NULL)
    out |>
      transmute(
        reporter = reporter_iso,
        partner  = partner_iso,
        year     = as.integer(ref_year),
        hs6      = as.character(cmd_code),
        value    = primary_value
      )
  }) |>
  bind_rows()

stopifnot(NROW(trade) > 0)

message("Observations: ", nrow(trade),
        " | reporters: ", n_distinct(trade$reporter),
        " | years: ", paste(range(trade$year), collapse = "-"))
```

## The metric

```{r windowing}
group_of <- function(p) {
  case_when(p == target ~ "russia",
            p %in% transit_hubs ~ "hub",
            p %in% control_partners ~ "control",
            TRUE ~ NA_character_)
}

# Each observation is assigned to the pre or post window of its own code.
windowed <- trade |>
  inner_join(select(code_windows, hs6, pre_from, pre_to, post_from, post_to,
                    pre_width, post_width, list),
             by = "hs6") |>
  mutate(
    group = group_of(partner),
    period = case_when(
      year >= pre_from  & year <= pre_to  ~ "pre",
      year >= post_from & year <= post_to ~ "post",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(group), !is.na(period))

# Annualising divides by the width of the window, not by the number of years
# actually observed. A year with no exports is a missing row in Comtrade
# rather than a zero, so dividing by observed years would quietly inflate
# the average for exactly the flows that stopped — which is the effect being
# measured.
annual <- windowed |>
  group_by(reporter, hs6, list, group, period) |>
  summarise(total = sum(value),
            width = first(if_else(period == "pre", pre_width, post_width)),
            .groups = "drop") |>
  mutate(annual_value = total / width) |>
  select(reporter, hs6, list, group, period, annual_value) |>
  pivot_wider(names_from = period, values_from = annual_value,
              values_fill = 0)
```

```{r metric}
# Control growth is what ordinary trade did over the same window for this
# reporter and this good. Hub growth is only interesting in excess of it.
control_growth <- annual |>
  filter(group == "control") |>
  transmute(reporter, hs6,
            control_factor = if_else(pre > 0, post / pre, NA_real_))

diversion <- annual |>
  filter(group %in% c("russia", "hub")) |>
  select(reporter, hs6, list, group, pre, post) |>
  pivot_wider(names_from = group, values_from = c(pre, post), values_fill = 0) |>
  left_join(control_growth, by = c("reporter", "hs6")) |>
  mutate(
    russia_lost = pmax(pre_russia - post_russia, 0),
    # What hub exports would have been had they merely followed ordinary
    # trade growth for this reporter and good.
    hub_expected = pre_hub * coalesce(control_factor, 1),
    hub_excess   = post_hub - hub_expected,
    # The headline: how much of the lost Russian trade re-appears as excess
    # exports to transit economies. Not a measure of guilt — a measure of
    # how much volume would need another explanation.
    diversion_ratio = if_else(russia_lost > 0, hub_excess / russia_lost, NA_real_)
  )

diversion |>
  filter(!is.na(diversion_ratio)) |>
  group_by(reporter) |>
  summarise(
    goods         = n(),
    lost_to_ru    = sum(russia_lost),
    excess_to_hub = sum(hub_excess),
    ratio         = sum(hub_excess) / sum(russia_lost),
    .groups = "drop"
  ) |>
  arrange(desc(ratio)) |>
  mutate(across(c(lost_to_ru, excess_to_hub), ~ round(.x / 1e6)),
         ratio = round(ratio, 2)) |>
  kable(caption = "Diversion ratio by country (values in EUR million per year)")
```

A ratio near zero means the trade simply stopped. A ratio near one means an
amount equivalent to the entire lost Russian trade re-appeared elsewhere,
beyond what ordinary growth explains.

```{r by_list}
# The CHP items are the ones subject to the Article 12gb duty. Splitting the
# ratio by list is the closest this data comes to testing whether an
# obligation aimed at exporters, rather than at goods, changed anything.
diversion |>
  filter(!is.na(diversion_ratio)) |>
  group_by(reporter, list) |>
  summarise(ratio = sum(hub_excess) / sum(russia_lost), .groups = "drop") |>
  pivot_wider(names_from = list, values_from = ratio) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = "Diversion ratio by code list")
```

## Validation

```{r validation}
# In July 2026 a Finnish district court convicted a company director of an
# aggravated regulation offence for exporting 164 trucks and trailers to
# Russia in 2022-2023, declared as sales to buyers registered in Turkey and
# Kazakhstan. The prosecution rested on investigative work, not on trade
# statistics. If this metric measures what it claims to, the pattern should
# be visible in Finnish exports of those goods to those two partners years
# before the judgment.
anchor_hs6 <- anchor_codes_tbl$hs6
anchor_partners <- c("TUR", "KAZ")

trade |>
  filter(reporter == "FIN", hs6 %in% anchor_hs6,
         partner %in% c(anchor_partners, target)) |>
  group_by(year, partner) |>
  summarise(value = sum(value) / 1e6, .groups = "drop") |>
  ggplot(aes(year, value, colour = partner)) +
  geom_line(linewidth = 1) +
  geom_point() +
  geom_vline(data = policy_events, aes(xintercept = year(date)),
             linetype = "dashed") +
  labs(x = NULL, y = "Exports, EUR million", colour = NULL,
       caption = paste("Finnish exports of trucks and trailers.",
                       "Dashed lines: statutory dates from Part 1.")) +
  theme_minimal()
```

This is the honest test of the whole approach. If the anchor pattern is
absent, the metric does not detect a case that is known to have happened,
and that has to be reported as such rather than explained away. If it is
present, an open-data method flagged, years earlier, a case the authorities
reached through investigation.

## Sensitivity

```{r sensitivity}
# The hub list is the most arguable choice in the report. Each country is
# removed in turn to see whether any single partner carries the result.
ratio_without <- function(dropped_hub) {
  hubs <- setdiff(transit_hubs, dropped_hub)
  windowed |>
    mutate(group = case_when(partner == target ~ "russia",
                             partner %in% hubs ~ "hub",
                             partner %in% control_partners ~ "control",
                             TRUE ~ NA_character_)) |>
    filter(!is.na(group)) |>
    group_by(reporter, group, period) |>
    summarise(total = sum(value),
              width = first(if_else(period == "pre", pre_width, post_width)),
              .groups = "drop") |>
    mutate(v = total / width) |>
    select(reporter, group, period, v) |>
    pivot_wider(names_from = c(group, period), values_from = v,
                values_fill = 0) |>
    transmute(reporter, dropped = dropped_hub,
              ratio = (hub_post - hub_pre) / pmax(russia_pre - russia_post, 1))
}

map_dfr(c("none", transit_hubs), ratio_without) |>
  pivot_wider(names_from = reporter, values_from = ratio) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = "Diversion ratio with each transit economy removed in turn")
```

## Limitations

- A diversion ratio is not evidence of circumvention. It measures volume
  that ordinary trade growth does not explain, which is a reason to look,
  not a finding of fact.
- The headline measurement uses the Common High Priority list, which is
  published at six digits and needs no truncation. The vehicle codes used
  for validation are truncated from eight digits to six, so restricted and
  unrestricted goods are counted together there.
- Baselines differ in length across codes where a code was restricted too
  recently for a full window, and a code with no post period at all is
  excluded. Both are reported in the output rather than averaged away.
- Mirror statistics disagree. What Finland records as an export to
  Kazakhstan and what Kazakhstan records as an import from Finland are
  frequently different numbers, and neither is authoritative.
- Transit trade is recorded inconsistently between countries, which is
  precisely the mechanism at issue in the Finnish case, where transit was
  not covered by the sanctions in force at the time.
- The transit-hub list is an assumption carried in from enforcement
  reporting. The sensitivity table shows how much rests on it.
- Goods routed through a country that is neither a hub on this list nor a
  control are invisible to this method entirely.
