---
title: "Where did the trade go?"
subtitle: "Measuring sanctions circumvention exposure across five Nordic countries"
author: "Kristian Vepsäläinen"
date: today
lang: en
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: false
execute:
  warning: false
  message: false
  cache: false
params:
  baseline_years: 3
  last_year: 2025
  regime_start: "2022-01-01"
  reporters: ["FIN", "SWE", "DNK", "NOR", "ISL"]
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(lubridate)
library(readr)
library(ggplot2)
library(knitr)

options(cli.dynamic = FALSE, cli.ansi = FALSE, cli.unicode = FALSE)

cache_dir <- here("data", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

cached <- function(name, expr) {
  path <- file.path(cache_dir, paste0(name, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  val <- force(expr)
  if (NROW(val) == 0) {
    warning("Empty result for '", name, "' — not cached.")
    return(val)
  }
  saveRDS(val, path)
  val
}
```

## The question

Sanctions on exports to Russia apply to the same goods across the EU and,
through aligned national measures, across Norway and Iceland. The rules are
common; exposure to their circumvention is not. A country's exposure
depends on what it exported to Russia before the restrictions, who its
neighbours are, and how much of that trade re-appeared shortly afterwards
in countries that had never bought it in any quantity.

This report measures that re-appearance in five Nordic countries, using
one metric applied identically to each.

## Method

Circumvention through a third country leaves a specific trace in customs
statistics. Exports of a restricted good to Russia fall to near zero, while
exports of the same good to a small number of transit economies rise far
beyond anything those economies could plausibly consume. The goods are not
consumed there; they are forwarded.

Three things make this measurable rather than suggestive:

**The goods are defined by the legislation, not by me.** The restricted
product codes are enumerated in the annexes of the EU sanctions
regulations, and the headline list is already published at the resolution
trade data uses. Part 1 extracts the codes and the date each one became
restricted; this part counts them. Choosing codes by hand would make any
finding an artefact of that choice.

**The comparison is against the country's own past, code by code.** Each
flow is measured against its own baseline in the years before that
particular code was restricted. A country that always traded heavily with
Central Asia is therefore not flagged for continuing to do so, and a code
restricted in 2025 is not judged against a 2022 boundary.

**A control group absorbs the rest.** Trade to partners with no plausible
onward route to Russia grew or shrank over the same period for ordinary
economic reasons. Differencing against that control removes the part of the
change that is simply the world economy.

## Part 1 — What the legislation defines

```{r codes}
# Two lists, because they are two different animals.
#
# The Common High Priority list is carried in the sanctions regulation
# itself, as Annex XL of Regulation 833/2014 (and Annex XXX of 765/2006 for
# Belarus). It holds 50 items in four tiers and — decisively here — it is
# expressed in six-digit HS codes, the same resolution as international
# trade data. It was drawn up jointly by the EU, US, UK and Japan
# specifically to support anti-circumvention work, so it is the list
# enforcement agencies themselves watch. No truncation is needed.
#
# The vehicle codes used for validation come from the industrial goods
# annex instead, at eight digits, and are truncated to HS6. That truncation
# is confined to the validation and never touches the headline measurement.
chp_codes <- tribble(
  ~tier,  ~hs6,     ~description,
  # Tier 1 — integrated circuits (microelectronics)
  "1",    "854231", "Processors, controllers, converters, logic circuits",
  "1",    "854232", "Memory integrated circuits",
  "1",    "854233", "Amplifier integrated circuits",
  "1",    "854239", "Other integrated circuits",
  # Tier 2 — wireless communications, satellite navigation, passive components
  "2",    "851762", "Voice/data transmission and switching apparatus",
  "2",    "852691", "Radio navigational aid apparatus",
  "2",    "853221", "Tantalum capacitors",
  "2",    "853224", "Ceramic dielectric multilayer capacitors",
  "2",    "854800", "Electrical machinery parts not elsewhere specified",
  # Tier 3.A — further electronic components
  "3.A",  "847150", "Processing units for data systems",
  "3.A",  "850440", "Static converters",
  "3.A",  "851769", "Apparatus for wired/wireless voice and data transmission",
  "3.A",  "852589", "Digital cameras and video recorders",
  "3.A",  "852910", "Aerials and aerial reflectors",
  "3.A",  "852990", "Parts for transmission/reception apparatus",
  "3.A",  "853669", "Plugs and sockets (<=1000V)",
  "3.A",  "853690", "Electrical switching apparatus (<=1000V)",
  "3.A",  "854110", "Diodes (non-LED)",
  "3.A",  "854121", "Transistors (<1W dissipation)",
  "3.A",  "854129", "Other transistors",
  "3.A",  "854130", "Thyristors, diacs, triacs",
  "3.A",  "854149", "Photosensitive semiconductor devices",
  "3.A",  "854151", "Semiconductor-based transducers",
  "3.A",  "854159", "Other semiconductor devices",
  "3.A",  "854160", "Mounted piezo-electric crystals",
  # Tier 3.B — mechanical and optical components
  "3.B",  "848210", "Ball bearings",
  "3.B",  "848220", "Tapered roller bearings",
  "3.B",  "848230", "Spherical roller bearings",
  "3.B",  "848250", "Cylindrical roller bearings",
  "3.B",  "880730", "Aircraft and helicopter parts",
  "3.B",  "901310", "Telescopic sights and periscopes",
  "3.B",  "901380", "Other optical devices",
  "3.B",  "901420", "Aeronautical/space navigation instruments",
  "3.B",  "901480", "Other navigational instruments",
  # Tier 4.A — manufacturing and test equipment for electronics
  "4.A",  "847180", "Data-processing machine units",
  "4.A",  "848610", "Boule/wafer manufacturing machines",
  "4.A",  "848620", "Semiconductor/integrated circuit manufacturing machines",
  "4.A",  "848640", "Specified semiconductor manufacturing apparatus",
  "4.A",  "853400", "Printed circuits",
  "4.A",  "854320", "Signal generators",
  "4.A",  "902750", "Optical radiation instruments (UV/visible/infrared)",
  "4.A",  "903020", "Oscilloscopes",
  "4.A",  "903032", "Multimeters with recording device",
  "4.A",  "903039", "Voltage/current/resistance measuring instruments",
  "4.A",  "903082", "Semiconductor wafer/device measurement instruments",
  # Tier 4.B — CNC machine tools (added February 2024)
  "4.B",  "845710", "Numerically controlled machining centres",
  "4.B",  "845811", "Numerically controlled horizontal lathes",
  "4.B",  "845891", "Numerically controlled lathes (non-horizontal)",
  "4.B",  "845961", "Numerically controlled metal milling machines",
  "4.B",  "846693", "Machine parts and accessories (headings 8456-8461)",
) |>
  mutate(regulation = "833/2014", annex = "XL",
         list = "CHP", cn8_truncated = FALSE)

stopifnot(nrow(chp_codes) == 50, !any(duplicated(chp_codes$hs6)))

# Plain-language names for the tiers. Six-digit codes carry no meaning to
# anyone who does not work with the nomenclature daily, and every table
# below is read by people who do not.
tier_labels <- c(
  "1"   = "Integrated circuits",
  "2"   = "Comms, navigation, passive components",
  "3.A" = "Electronic components",
  "3.B" = "Bearings, optics, navigation instruments",
  "4.A" = "Electronics manufacturing and test equipment",
  "4.B" = "CNC machine tools"
)

chp_codes <- mutate(chp_codes, category = unname(tier_labels[tier]))
stopifnot(!any(is.na(chp_codes$category)))

# The vehicle codes carry no annex number here on purpose. Which annex they
# sit in has changed across packages, and nothing downstream depends on it —
# the restriction date comes from the amending act, not from the annex
# label. The CELEX of the act that imposed each one is recovered below and
# is the citable reference.
anchor_codes_tbl <- tribble(
  ~hs6,     ~description,
  "870421", "Motor vehicles for goods, diesel, <=5t",
  "870422", "Motor vehicles for goods, diesel, 5-20t",
  "871620", "Trailers and semi-trailers",
) |>
  mutate(tier = NA_character_, regulation = "833/2014",
         annex = NA_character_, list = "Industrial", cn8_truncated = TRUE,
         category = "Vehicles and trailers")

target_codes <- unique(c(chp_codes$hs6, anchor_codes_tbl$hs6))

# One lookup for anything that needs to put a human-readable name next to a
# code, so that the review file and every table describe a good the same way.
code_meta <- bind_rows(chp_codes, anchor_codes_tbl) |>
  select(hs6, tier, category, description, list)
```

### When each code actually became restricted

The date a code became restricted is deliberately not taken from Annex XL.
Annex XL is where the CHP list is *carried*, but the annex itself was only
created by the 14th package in June 2024, to give Article 12gb something to
point at. The export bans on these goods are older and arrived in different
packages — most of the electronics under the advanced-technology annex
during 2022, the CNC machine tools later. Using the date Annex XL appeared
would push years of already-restricted trade into the "before" window and
quietly destroy the comparison.

So the dates are derived from the regulation itself: every act amending
Regulation 833/2014, in order, and the first one in which each code appears.
The acts are found two ways — through the `eurlex` package's query builder
and through the amendment relation in the underlying ontology — and the two
are reconciled before anything is built on them.

```{r eurlex_helpers}
library(eurlex)

# eurlex returns different column names depending on which include_ options
# were set and on the package version, so columns are looked up by a list of
# plausible names rather than assumed.
pick <- function(df, ..., default = NA_character_) {
  nms <- intersect(c(...), names(df))
  if (length(nms) == 0) return(rep(default, nrow(df)))
  df[[nms[1]]]
}

# Entry into force is the operative date. Where Cellar has not recorded one,
# the document date stands in and the substitution is flagged, because the
# two can differ by days and that matters at a year boundary.
normalise_acts <- function(df, route) {
  tibble(
    route      = route,
    work       = pick(df, "work"),
    celex      = pick(df, "celex"),
    title      = pick(df, "title"),
    date_force = suppressWarnings(as.Date(pick(df, "dateforce", "date_force"))),
    date_doc   = suppressWarnings(as.Date(pick(df, "date", "date_doc")))
  ) |>
    mutate(
      effective        = coalesce(date_force, date_doc),
      date_is_fallback = is.na(date_force)
    ) |>
    filter(!is.na(celex), !is.na(effective)) |>
    distinct(celex, .keep_all = TRUE)
}

# A route that yields nothing must still return the same shape, so that the
# reconciliation below compares two tables rather than a table and a NULL.
empty_acts <- function() {
  tibble(route = character(), work = character(), celex = character(),
         title = character(), date_force = as.Date(character()),
         date_doc = as.Date(character()), effective = as.Date(character()),
         date_is_fallback = logical())
}
```

```{r eurlex_route_a}
# Route A — the package's own query builder.
#
# Every act amending the regulation carries it in the title ("... amending
# Regulation (EU) No 833/2014"), so the whole step fits inside the eurlex
# API. The advantage over hand-written SPARQL is that the selection is
# readable without knowing SPARQL, and it survives changes to the Cellar
# ontology, which the raw route does not.
# elx_make_query gained include_title and date_from after the version
# currently on CRAN, so the arguments are passed only where the installed
# version accepts them. Both are load-bearing here: without a title there is
# nothing to match on, and without a date bound the query asks Cellar for
# every regulation ever adopted. So where they are missing this route is
# skipped outright rather than run in a degraded form that would look like
# a result.
elx_args <- names(formals(elx_make_query))
route_a_args <- c("include_title", "date_from")
route_a_ok <- all(route_a_args %in% elx_args)

acts_a <- if (!route_a_ok) {
  message("eurlex ", as.character(utils::packageVersion("eurlex")),
          " lacks ", paste(setdiff(route_a_args, elx_args), collapse = ", "),
          "; Route A skipped and the cross-check is unavailable. ",
          "remotes::install_github('michalovadek/eurlex') restores it.")
  empty_acts()
} else {
  amend_query_pkg <- elx_make_query(
    resource_type      = "regulation",
    include_celex      = TRUE,
    include_date       = TRUE,
    include_date_force = TRUE,
    include_title      = TRUE,
    date_from          = "2022-01-01",
    order              = TRUE
  )

  cached("eurlex_acts_route_a", {
    elx_run_query(amend_query_pkg) |>
      as_tibble() |>
      rename_with(tolower) |>
      filter(str_detect(coalesce(title, ""), "833/2014")) |>
      normalise_acts("package")
  })
}

message("Route A (elx_make_query + title filter): ", nrow(acts_a), " acts")
```

```{r eurlex_route_b}
# Route B — the amendment relation itself.
#
# elx_make_query has no argument for it: the include_ options cover dates,
# legal bases, citations, EuroVoc and court metadata, but the only relation
# between documents it exposes is cdm:work_cites_work. Amendment is a
# different and stricter relation, so this one goes in as SPARQL and runs
# through elx_run_query like any other query.
#
# It is kept as a check on Route A rather than as the primary source: a
# title filter is a proxy for "amends", and two independent routes agreeing
# is worth more than either one on its own. (If the relation helpers in the
# development version on GitHub land on CRAN, this block becomes a one-line
# call and the SPARQL can go.)
amend_query_sparql <- '
PREFIX cdm: <http://publications.europa.eu/ontology/cdm#>
PREFIX xsd: <http://www.w3.org/2001/XMLSchema#>
SELECT DISTINCT ?work ?celex ?dateforce ?date
WHERE {
  ?work cdm:resource_legal_amends_resource_legal ?amended .
  ?amended cdm:resource_legal_id_celex "32014R0833"^^xsd:string .
  OPTIONAL { ?work cdm:resource_legal_id_celex ?celex . }
  OPTIONAL { ?work cdm:resource_legal_date_entry-into-force ?dateforce . }
  OPTIONAL { ?work cdm:work_date_document ?date . }
}
ORDER BY ?dateforce'

acts_b <- cached("eurlex_acts_route_b", {
  res <- tryCatch(elx_run_query(amend_query_sparql),
                  error = function(e) {
                    warning("Route B failed: ", conditionMessage(e))
                    NULL
                  })
  if (is.null(res) || NROW(res) == 0) {
    empty_acts()
  } else {
    res |> as_tibble() |> rename_with(tolower) |> normalise_acts("sparql")
  }
})

message("Route B (amends relation): ", nrow(acts_b), " acts")
```

```{r eurlex_reconcile}
# Where the two routes disagree is worth knowing before anything is built on
# top of them. Only in Route A usually means a related act that mentions the
# regulation without amending it; only in Route B means an amendment whose
# title is phrased differently, or one dated before the 2022 cut-off.
only_a <- setdiff(acts_a$celex, acts_b$celex)
only_b <- setdiff(acts_b$celex, acts_a$celex)

tibble(
  `Route A (title)` = if (route_a_ok) as.character(nrow(acts_a)) else "not run",
  `Route B (amends)` = as.character(nrow(acts_b)),
  `In both` = length(intersect(acts_a$celex, acts_b$celex)),
  `Only A` = length(only_a),
  `Only B` = length(only_b)
) |>
  kable(caption = "Agreement between the two EUR-Lex routes")

# The union is used, not the intersection. Missing an act that restricted a
# code would silently date that code too late, which biases the measurement;
# including an act that merely mentions the regulation costs one extra text
# fetch and nothing else, because a code is only dated from an act whose
# text actually contains it.
amendments_all <- bind_rows(acts_a, acts_b) |>
  arrange(effective) |>
  distinct(celex, .keep_all = TRUE)

# Regulation 833/2014 dates from the Crimea measures, and the amendment
# relation reaches back to 2014. Those older acts have to go, and not for
# tidiness: a code first restricted in 2014 would be dated to 2014, its
# "post" window would run from 2015 to 2025, and the 2022 rupture this
# report is about would be averaged into a decade of ordinary trade.
#
# So the measurement is scoped to the post-invasion regime. A code carried
# over from the Crimea-era measures and never re-listed after 2022 appears
# in no act in scope and drops out, which is correct: it is outside the
# regime being measured, not a gap in the data.
amendments <- filter(amendments_all,
                     effective >= as.Date(params$regime_start))

tibble(
  `Acts found` = nrow(amendments_all),
  `Before regime start` = nrow(amendments_all) - nrow(amendments),
  `Carried forward` = nrow(amendments)
) |>
  kable(caption = paste("Scoping to the regime beginning",
                        params$regime_start))

if (nrow(amendments) == 0) {
  stop("Neither EUR-Lex route returned any amending act. Route A needs a ",
       "newer eurlex; Route B needs the ontology predicate ",
       "cdm:resource_legal_amends_resource_legal to still be current.")
}

message("Acts carried forward: ", nrow(amendments),
        " (", sum(amendments$date_is_fallback), " using document date)")
```

```{r eurlex_text}
text_dir <- here("data", "eurlex")
dir.create(text_dir, recursive = TRUE, showWarnings = FALSE)

# elx_fetch_data needs a full resolvable URL. The ?work values coming back
# from the endpoint arrive as bare Cellar UUIDs, which curl then tries to
# resolve as a hostname — so the URL is built here rather than taken from
# the query result. CELEX is used in preference to the Cellar UUID because
# it is stable, it is the identifier a lawyer would cite, and it makes the
# provenance column in the review file readable instead of a hex string.
celex_url <- function(celex) {
  paste0("http://publications.europa.eu/resource/celex/", celex)
}

# 32024R1745 -> "Reg (EU) 2024/1745". A CELEX number is precise and
# unreadable; the short form is how the act is referred to in practice, and
# both appear in the review file so nothing is lost.
celex_short <- function(celex) {
  sub("^3(\\d{4})R(\\d{4})$", "Reg (EU) \\1/\\2", celex)
}

# Full text is fetched once per act and kept on disk. Cellar is a shared
# public service and there is no reason to ask it for the same document
# twice.
fetch_act_text <- function(celex) {
  path <- file.path(text_dir, paste0(celex, ".txt"))
  if (file.exists(path)) return(read_file(path))
  Sys.sleep(1)
  txt <- tryCatch(
    elx_fetch_data(url = celex_url(celex), type = "text", language_1 = "en"),
    error = function(e) {
      warning(celex, ": ", conditionMessage(e))
      NA_character_
    }
  )
  txt <- paste(stats::na.omit(txt), collapse = "\n")
  if (!nzchar(txt)) return(NA_character_)
  write_file(txt, path)
  txt
}

# CN codes in annex text appear as "8542 31 00", "8542 31" or "85423100",
# often prefixed "ex". Parsing the annex tables themselves is not worth it:
# they flatten unpredictably in the text rendering and the layout changes
# between packages. Instead every digit group that could be a CN code is
# extracted and then intersected with the codes actually being looked for.
# The regex is noisy; the intersection is not, because a six-digit string
# that happens to match one of 53 target codes by accident is rare and shows
# up as an implausible date in the review file.
extract_hs6 <- function(txt) {
  if (is.na(txt) || !nzchar(txt)) return(character(0))
  str_extract_all(txt, "\\b\\d{4}\\s?\\d{2}(\\s?\\d{2})?\\b")[[1]] |>
    str_remove_all("\\s") |>
    str_sub(1, 6) |>
    unique()
}

code_appearances <- cached("eurlex_code_appearances", {
  map_dfr(amendments$celex, function(celex) {
    txt <- fetch_act_text(celex)
    found <- intersect(extract_hs6(txt), target_codes)
    if (length(found) == 0) return(NULL)
    tibble(celex = celex, hs6 = found)
  })
})

fetched <- length(list.files(text_dir, pattern = "\\.txt$"))
message("Text retrieved for ", fetched, " of ", nrow(amendments), " acts | ",
        "codes located: ", n_distinct(code_appearances$hs6),
        " of ", length(target_codes))

# A near-total fetch failure produces an empty appearance table, which would
# otherwise surface much later as "no code has a date" and look like a
# regulatory finding rather than a broken download.
if (fetched < 0.5 * nrow(amendments)) {
  stop("Text could only be retrieved for ", fetched, " of ", nrow(amendments),
       " acts. Check that ", celex_url(amendments$celex[1]),
       " resolves before reading anything below.")
}
```

```{r eurlex_effective}
code_effective_auto <- code_appearances |>
  left_join(select(amendments, celex, effective), by = "celex") |>
  group_by(hs6) |>
  slice_min(effective, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(hs6, effective, imposed_by = celex, source = "eurlex")

# First appearance is a good approximation, not a fact. A code can show up
# in an amending act because an entry was reworded, moved between annexes,
# or restricted for import rather than export. Rows verified against the
# regulation text go here and override the extraction.
code_effective_manual <- tribble(
  ~hs6, ~effective, ~imposed_by,
  # "854231", "2022-04-08", "32022R0576 (5th package), Annex VII",
)

code_effective_manual <- if (nrow(code_effective_manual) > 0) {
  mutate(code_effective_manual,
         effective = as.Date(effective), source = "manual")
} else {
  tibble(hs6 = character(), effective = as.Date(character()),
         imposed_by = character(), source = character())
}

code_effective <- code_effective_auto |>
  filter(!hs6 %in% code_effective_manual$hs6) |>
  bind_rows(code_effective_manual)

# Written out so the extraction can be checked against the regulation text
# rather than trusted. This is the step that turns a scrape into evidence,
# and it is the file to open first when a result looks surprising.
review_dir <- here("data", "derived")
dir.create(review_dir, recursive = TRUE, showWarnings = FALSE)
code_effective |>
  left_join(code_meta, by = "hs6") |>
  mutate(act = celex_short(imposed_by), act_url = celex_url(imposed_by)) |>
  select(hs6, category, description, tier, list, effective, act, imposed_by,
         act_url, source) |>
  arrange(effective, hs6) |>
  write_csv(file.path(review_dir, "code_effective_for_review.csv"))

not_found <- setdiff(target_codes, code_effective$hs6)
if (length(not_found) > 0) {
  message(length(not_found), " code(s) appear in no act in scope and will ",
          "be excluded: ",
          code_meta |>
            filter(hs6 %in% not_found) |>
            mutate(lab = paste0(hs6, " (", description, ")")) |>
            pull(lab) |>
            paste(collapse = "; "))
}

# Coverage by category rather than a second list of dates. What matters
# before reading any result is whether a whole category failed to date,
# because that would remove a kind of good from the comparison entirely
# rather than thinning it evenly.
code_meta |>
  left_join(select(code_effective, hs6, effective), by = "hs6") |>
  group_by(Category = category) |>
  summarise(
    Codes = n(),
    Dated = sum(!is.na(effective)),
    `First restricted` = suppressWarnings(min(effective, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(`First restricted` = if_else(is.finite(`First restricted`),
                                      `First restricted`, as.Date(NA))) |>
  arrange(Category) |>
  kable(caption = "Date coverage by product category")
```

```{r assemble_codes}
restricted_codes <- bind_rows(chp_codes, anchor_codes_tbl) |>
  left_join(code_effective, by = "hs6")

# A code without an effective date has no baseline, so it is dropped rather
# than compared against arbitrary years. A handful of misses is a gap in the
# extraction; losing most of the list means the SPARQL or the regex is
# broken, and the run stops rather than reporting a confident number built
# on a tenth of the goods.
coverage <- mean(!is.na(restricted_codes$effective))
if (coverage < 0.7) {
  stop(sprintf(paste(
    "Only %.0f%% of codes (%d of %d) have a restriction date.",
    "Check code_appearances and the EUR-Lex query before going further;",
    "nothing below this point will run."),
    100 * coverage, sum(!is.na(restricted_codes$effective)),
    nrow(restricted_codes)))
}
if (coverage < 1) {
  message(sprintf("%.0f%% of codes dated; %d dropped for lack of a date.",
                  100 * coverage, sum(is.na(restricted_codes$effective))))
}

restricted_codes <- filter(restricted_codes, !is.na(effective))

restricted_codes |>
  count(list, cn8_truncated, name = "codes") |>
  kable(caption = "Restricted goods carried into the measurement")
```

Codes carry their own start date because the restrictions did not arrive at
once. Twenty-one sanctions packages had been adopted by August 2026, each
one amending the annexes: entries added, codes refined, thresholds revised.
A code restricted from 2023 and a code restricted from 2025 cannot share a
single before-and-after boundary without mixing restricted and unrestricted
years inside the same comparison.

```{r windows}
# Each code gets its own baseline and its own post period. The year the
# restriction takes effect is dropped: it is part restricted and part not,
# and assigning it to either side would bias the comparison in a direction
# that depends on the month of adoption.
code_windows <- restricted_codes |>
  mutate(
    eff_year  = year(effective),
    pre_from  = eff_year - params$baseline_years,
    pre_to    = eff_year - 1,
    post_from = eff_year + 1,
    post_to   = params$last_year,
    pre_width  = pre_to - pre_from + 1,
    post_width = post_to - post_from + 1
  ) |>
  filter(post_width >= 1)

dropped <- nrow(restricted_codes) - nrow(code_windows)
if (dropped > 0) {
  message(dropped, " code(s) restricted too recently to have a post period ",
          "and excluded.")
}

code_windows |>
  count(effective, imposed_by, category, name = "codes") |>
  arrange(effective, desc(codes)) |>
  transmute(
    Date     = effective,
    Act      = celex_short(imposed_by),
    Category = category,
    Codes    = codes,
    Baseline = paste0(year(effective) - params$baseline_years, "–",
                      year(effective) - 1)
  ) |>
  kable(caption = paste(
    "What each act restricted, and the baseline years each code is",
    "measured against"))
```

```{r policy_events}
# Part 1's second output: obligations whose effect Part 2 can test.
# Article 12gb of Regulation 833/2014 is the interesting one, because it
# does not restrict a good — it obliges exporters to take active steps
# against circumvention of the CHP items specifically. If that obligation
# bites, its effect should appear in CHP flows and not in the others, which
# is a sharper prediction than "sanctions reduce trade".
# Each row is either a legal act, identified by CELEX and therefore
# checkable against the text, or a dated context marker with no instrument
# behind it. The two are kept apart so that nothing downstream cites an
# event as though it were a provision.
#
# The date an obligation starts to apply is not the date its act entered
# into force. Article 12gb was enacted by the June 2024 package and applies
# from 26 December 2024, so the date and the CELEX below deliberately
# disagree.
policy_events <- tribble(
  ~date,        ~event,                                              ~celex,
  "2022-02-24", "Full-scale invasion; sectoral packages begin",       NA_character_,
  "2024-12-26", "Art. 12gb anti-circumvention duty applies (Russia)", "32024R1745",
) |>
  mutate(
    date   = as.Date(date),
    kind   = if_else(is.na(celex), "context", "legal act"),
    source = if_else(is.na(celex), NA_character_, celex_url(celex))
  )

# An act cited here that the EUR-Lex query never returned means either the
# attribution is wrong or the scoping dropped something it should not have.
# Either way it is worth knowing before the date is used as a boundary.
unverified <- setdiff(stats::na.omit(policy_events$celex), amendments$celex)
if (length(unverified) > 0) {
  warning("policy_events cites acts not among those retrieved: ",
          paste(unverified, collapse = ", "))
}

policy_events |>
  transmute(Date = date, Event = event, Kind = kind,
            Act = coalesce(celex_short(celex), "—")) |>
  kable(caption = "Obligations tested in Part 2")
```

## Part 2 — What the trade data shows

### Partner groups

```{r partners}
# Transit economies are the ones repeatedly identified in EU and national
# enforcement reporting as onward routes to Russia. The list is a stated
# assumption, not a finding, and the results are re-run without each one in
# the sensitivity section.
transit_hubs <- c("ARM", "KAZ", "KGZ", "GEO", "UZB", "TUR", "ARE", "SRB")

# Controls: economies of comparable income and distance with no land or
# practical onward route to Russia. They absorb ordinary trade growth.
control_partners <- c("MAR", "TUN", "JOR", "THA", "MYS", "CHL", "ZAF")

target <- "RUS"
```

### Data

```{r trade_data}
# Running chunks by hand makes it easy to arrive here after an earlier chunk
# stopped, and "object not found" points at the wrong place. This names the
# chunk that is actually missing.
if (!exists("code_windows")) {
  stop("code_windows does not exist — the 'windows' chunk has not run. ",
       "Something before it stopped; check the assemble_codes coverage ",
       "check and the EUR-Lex chunks above.")
}

# UN Comtrade is used for all five reporters rather than Eurostat, because
# Norway and Iceland are outside Comext and a cross-country comparison built
# from two different collection systems would confound the measurement with
# the measurement apparatus. Comext at CN8 is the better source for the
# three EU members alone and belongs in the robustness check.
#
# comtradr::set_primary_comtrade_key() needs a free API key.
library(comtradr)

years <- seq(min(code_windows$pre_from), params$last_year)

fetch_flow <- function(reporter, year) {
  cached(sprintf("trade_%s_%d", reporter, year), {
    ct_get_data(
      reporter       = reporter,
      partner        = c(target, transit_hubs, control_partners),
      commodity_code = unique(code_windows$hs6),
      start_date     = year,
      end_date       = year,
      flow_direction = "export",
      freq           = "A"
    )
  })
}

trade <- expand_grid(reporter = params$reporters, year = years) |>
  pmap(function(reporter, year) {
    out <- tryCatch(fetch_flow(reporter, year), error = function(e) {
      warning(reporter, " ", year, ": ", conditionMessage(e))
      NULL
    })
    if (is.null(out) || NROW(out) == 0) return(NULL)
    out |>
      transmute(
        reporter = reporter_iso,
        partner  = partner_iso,
        year     = as.integer(ref_year),
        hs6      = as.character(cmd_code),
        value    = primary_value
      )
  }) |>
  bind_rows()

stopifnot(NROW(trade) > 0)

message("Observations: ", nrow(trade),
        " | reporters: ", n_distinct(trade$reporter),
        " | years: ", paste(range(trade$year), collapse = "-"))
```

## The metric

```{r windowing}
group_of <- function(p) {
  case_when(p == target ~ "russia",
            p %in% transit_hubs ~ "hub",
            p %in% control_partners ~ "control",
            TRUE ~ NA_character_)
}

# Each observation is assigned to the pre or post window of its own code.
windowed <- trade |>
  inner_join(select(code_windows, hs6, pre_from, pre_to, post_from, post_to,
                    pre_width, post_width, list),
             by = "hs6") |>
  mutate(
    group = group_of(partner),
    period = case_when(
      year >= pre_from  & year <= pre_to  ~ "pre",
      year >= post_from & year <= post_to ~ "post",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(group), !is.na(period))

# Annualising divides by the width of the window, not by the number of years
# actually observed. A year with no exports is a missing row in Comtrade
# rather than a zero, so dividing by observed years would quietly inflate
# the average for exactly the flows that stopped — which is the effect being
# measured.
annual <- windowed |>
  group_by(reporter, hs6, list, group, period) |>
  summarise(total = sum(value),
            width = first(if_else(period == "pre", pre_width, post_width)),
            .groups = "drop") |>
  mutate(annual_value = total / width) |>
  select(reporter, hs6, list, group, period, annual_value) |>
  pivot_wider(names_from = period, values_from = annual_value,
              values_fill = 0)
```

```{r metric}
# Control growth is what ordinary trade did over the same window for this
# reporter and this good. Hub growth is only interesting in excess of it.
control_growth <- annual |>
  filter(group == "control") |>
  transmute(reporter, hs6,
            control_factor = if_else(pre > 0, post / pre, NA_real_))

diversion <- annual |>
  filter(group %in% c("russia", "hub")) |>
  select(reporter, hs6, list, group, pre, post) |>
  pivot_wider(names_from = group, values_from = c(pre, post), values_fill = 0) |>
  left_join(control_growth, by = c("reporter", "hs6")) |>
  mutate(
    russia_lost = pmax(pre_russia - post_russia, 0),
    # What hub exports would have been had they merely followed ordinary
    # trade growth for this reporter and good.
    hub_expected = pre_hub * coalesce(control_factor, 1),
    hub_excess   = post_hub - hub_expected,
    # The headline: how much of the lost Russian trade re-appears as excess
    # exports to transit economies. Not a measure of guilt — a measure of
    # how much volume would need another explanation.
    diversion_ratio = if_else(russia_lost > 0, hub_excess / russia_lost, NA_real_)
  )

diversion |>
  filter(!is.na(diversion_ratio)) |>
  group_by(reporter) |>
  summarise(
    goods         = n(),
    lost_to_ru    = sum(russia_lost),
    excess_to_hub = sum(hub_excess),
    ratio         = sum(hub_excess) / sum(russia_lost),
    .groups = "drop"
  ) |>
  arrange(desc(ratio)) |>
  mutate(across(c(lost_to_ru, excess_to_hub), ~ round(.x / 1e6)),
         ratio = round(ratio, 2)) |>
  kable(caption = "Diversion ratio by country (values in EUR million per year)")
```

A ratio near zero means the trade simply stopped. A ratio near one means an
amount equivalent to the entire lost Russian trade re-appeared elsewhere,
beyond what ordinary growth explains.

```{r by_list}
# The CHP items are the ones subject to the Article 12gb duty. Splitting the
# ratio by list is the closest this data comes to testing whether an
# obligation aimed at exporters, rather than at goods, changed anything.
diversion |>
  filter(!is.na(diversion_ratio)) |>
  group_by(reporter, list) |>
  summarise(ratio = sum(hub_excess) / sum(russia_lost), .groups = "drop") |>
  pivot_wider(names_from = list, values_from = ratio) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = "Diversion ratio by code list")
```

## Validation

```{r validation}
# In July 2026 a Finnish district court convicted a company director of an
# aggravated regulation offence for exporting 164 trucks and trailers to
# Russia in 2022-2023, declared as sales to buyers registered in Turkey and
# Kazakhstan. The prosecution rested on investigative work, not on trade
# statistics. If this metric measures what it claims to, the pattern should
# be visible in Finnish exports of those goods to those two partners years
# before the judgment.
anchor_hs6 <- anchor_codes_tbl$hs6
anchor_partners <- c("TUR", "KAZ")

trade |>
  filter(reporter == "FIN", hs6 %in% anchor_hs6,
         partner %in% c(anchor_partners, target)) |>
  group_by(year, partner) |>
  summarise(value = sum(value) / 1e6, .groups = "drop") |>
  ggplot(aes(year, value, colour = partner)) +
  geom_line(linewidth = 1) +
  geom_point() +
  geom_vline(data = policy_events, aes(xintercept = year(date)),
             linetype = "dashed") +
  labs(x = NULL, y = "Exports, EUR million", colour = NULL,
       caption = paste("Finnish exports of trucks and trailers.",
                       "Dashed lines: statutory dates from Part 1.")) +
  theme_minimal()
```

This is the honest test of the whole approach. If the anchor pattern is
absent, the metric does not detect a case that is known to have happened,
and that has to be reported as such rather than explained away. If it is
present, an open-data method flagged, years earlier, a case the authorities
reached through investigation.

## Sensitivity

```{r sensitivity}
# The hub list is the most arguable choice in the report. Each country is
# removed in turn to see whether any single partner carries the result.
ratio_without <- function(dropped_hub) {
  hubs <- setdiff(transit_hubs, dropped_hub)
  windowed |>
    mutate(group = case_when(partner == target ~ "russia",
                             partner %in% hubs ~ "hub",
                             partner %in% control_partners ~ "control",
                             TRUE ~ NA_character_)) |>
    filter(!is.na(group)) |>
    group_by(reporter, group, period) |>
    summarise(total = sum(value),
              width = first(if_else(period == "pre", pre_width, post_width)),
              .groups = "drop") |>
    mutate(v = total / width) |>
    select(reporter, group, period, v) |>
    pivot_wider(names_from = c(group, period), values_from = v,
                values_fill = 0) |>
    transmute(reporter, dropped = dropped_hub,
              ratio = (hub_post - hub_pre) / pmax(russia_pre - russia_post, 1))
}

map_dfr(c("none", transit_hubs), ratio_without) |>
  pivot_wider(names_from = reporter, values_from = ratio) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = "Diversion ratio with each transit economy removed in turn")
```

## Limitations

- A diversion ratio is not evidence of circumvention. It measures volume
  that ordinary trade growth does not explain, which is a reason to look,
  not a finding of fact.
- The headline measurement uses the Common High Priority list, which is
  published at six digits and needs no truncation. The vehicle codes used
  for validation are truncated from eight digits to six, so restricted and
  unrestricted goods are counted together there.
- Baselines differ in length across codes where a code was restricted too
  recently for a full window, and a code with no post period at all is
  excluded. Both are reported in the output rather than averaged away.
- Restriction dates are derived by finding the first amending act whose
  text contains each code. A code can appear in an act because an entry was
  reworded or moved between annexes rather than newly restricted, so the
  dates are an approximation until checked. The extraction is written to
  `data/derived/code_effective_for_review.csv` for exactly that check, and
  verified rows override it.
- Entry-into-force dates come from Cellar. Where Cellar records none, the
  document date is used instead and the substitution is counted in the
  output, because the two can straddle a year boundary.
- The set of amending acts is the union of a title match and the ontology's
  amendment relation. Neither is authoritative on its own: the title filter
  is a proxy, and the relation depends on a predicate name that the ontology
  can revise. The reconciliation table reports how far apart they were.
- Only acts from the start of the post-invasion regime are in scope. A good
  restricted under the 2014 Crimea measures and not re-listed since is
  therefore absent from the analysis, even though it remains restricted.
- Mirror statistics disagree. What Finland records as an export to
  Kazakhstan and what Kazakhstan records as an import from Finland are
  frequently different numbers, and neither is authoritative.
- Transit trade is recorded inconsistently between countries, which is
  precisely the mechanism at issue in the Finnish case, where transit was
  not covered by the sanctions in force at the time.
- The transit-hub list is an assumption carried in from enforcement
  reporting. The sensitivity table shows how much rests on it.
- Goods routed through a country that is neither a hub on this list nor a
  control are invisible to this method entirely.
