---
title: "Who can measure whether beneficial ownership rules are followed?"
subtitle: "Four Nordic countries, one directive, and the only register an outsider can count"
author: "Kristian Vepsäläinen"
date: today
lang: en
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: false
execute:
  warning: false
  message: false
  cache: false
params:
  prh_base: "https://avoindata.prh.fi"
  notice_code: "ESR"
  start_year: 2015
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(lubridate)
library(httr2)
library(jsonlite)
library(ggplot2)
library(knitr)

options(cli.dynamic = FALSE, cli.ansi = FALSE, cli.unicode = FALSE)

cache_dir <- here("data", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

cached <- function(name, expr) {
  path <- file.path(cache_dir, paste0(name, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  val <- force(expr)
  n <- if (is.null(val)) 0 else NROW(val)
  if (n == 0) {
    warning("Empty result for '", name, "' — not cached.")
    return(val)
  }
  saveRDS(val, path)
  val
}
```

## The question

Every EU and EEA state requires companies to register who ultimately owns
or controls them. The obligation is near-identical across the Nordics
because it comes from the same directives. What differs is the penalty
ladder each country attached to it, and how quickly that ladder tightens
when companies do not comply.

Part 1 establishes, from the legislation itself, what each country says
should happen to a company that fails to file. Part 2 measures whether it
happened — in the one country where an outsider still can.

## What an outsider can actually see

The 2022 CJEU judgment in WM and Sovim ended public access to beneficial
ownership registers, and each country closed differently. The result is
that the underlying ownership data is now out of reach almost everywhere.
Finland is the exception in one narrow but decisive respect: the *fact and
date of a filing* is register metadata rather than beneficial ownership
data, and it remains open.

| Country | Ownership data | Filing metadata | Usable here |
|---|---|---|---|
| Finland | Closed | Open API, notice code `ESR`, from Nov 2014 | Yes |
| Sweden | Restricted to authorities and obliged entities; tightened July 2026 | No | No |
| Denmark | Closed to the public since Sept 2025 | No | No |
| Norway | Restricted to Norwegian users with legitimate interest | No | No |
| Iceland | Public on the registry website, no API | — | Scraping only |

This table is a finding, not a preamble. Four countries implement the same
obligation, and in exactly one of them can a third party count how well it
is being met. Everywhere else, the only way to assess a counterparty's
compliance is to buy the answer from a commercial provider.

## Part 1 — What the law says should happen

*This section is produced with the lexverse packages (finlex, swelex,
danlex, eurlex) and states, per country: who must file, by when, and the
ordered sequence of consequences for not filing. Its output is the table
below, which Part 2 then tests against the data.*

```{r policy_dates}
# The dates and sanction steps below are the output of Part 1 and the input
# to Part 2. Every row must be traceable to a provision — this is the seam
# where the legal reading becomes a testable prediction, so an unsourced
# row here would quietly become an unsourced claim in the results.
policy_events <- tribble(
  ~date,         ~country,  ~event,                              ~source,
  "2019-07-01",  "FI",      "Filing obligation enters into force", "TODO: provision",
  "2020-07-01",  "FI",      "Deadline for existing companies",     "TODO: provision",
  "2025-07-31",  "NO",      "Mandatory filing deadline, daily fines follow", "TODO",
) |>
  mutate(date = as.Date(date))

policy_events |> kable(caption = "Regulatory timeline feeding the measurement")
```

If Part 1 is right that the sanction ladder escalates at specific dates,
filing behaviour should respond at those dates and not at arbitrary ones.
That is the prediction Part 2 tests.

## Part 2 — What the data shows

### Finding the endpoints

```{r api_discovery}
# The exact REST paths are read from the service's own OpenAPI description
# rather than hard-coded, so that a change at PRH surfaces as a readable
# list instead of a 404 deep in a loop.
spec_candidates <- c(
  paste0(params$prh_base, "/opendata-registerednotices-api/v3/api-docs"),
  paste0(params$prh_base, "/fi/krek/v3/api-docs"),
  paste0(params$prh_base, "/krek/v3/api-docs")
)

spec <- NULL
for (u in spec_candidates) {
  spec <- tryCatch(
    request(u) |> req_perform() |> resp_body_json(),
    error = function(e) NULL
  )
  if (!is.null(spec)) {
    message("OpenAPI spec found at: ", u)
    break
  }
}

if (is.null(spec)) {
  message("No spec reachable. Open ", params$prh_base,
          "/fi/krek/swagger-ui in a browser and set notices_path below by hand.")
} else {
  message("Endpoints: ", paste(names(spec$paths), collapse = ", "))
}
```

```{r api_layer}
# Everything about the API lives here. If PRH changes a path or a parameter
# name, this is the only block that needs editing.
notices_path <- "/opendata-registerednotices-api/v3/notices"

fetch_notices_page <- function(start_date, end_date, page = 1) {
  request(paste0(params$prh_base, notices_path)) |>
    req_url_query(
      registrationDateStart = format(start_date, "%Y-%m-%d"),
      registrationDateEnd   = format(end_date, "%Y-%m-%d"),
      page                  = page
    ) |>
    req_retry(max_tries = 4, backoff = ~ 5 * .x) |>
    req_perform() |>
    resp_body_json(simplifyVector = TRUE)
}

# Notices are fetched a month at a time and cached per month. A decade of
# daily register activity is too much for one request, and monthly chunks
# make a partial failure cheap to resume.
fetch_month <- function(month_start) {
  key <- paste0("notices_", format(month_start, "%Y_%m"))
  cached(key, {
    month_end <- ceiling_date(month_start, "month") - days(1)
    pages <- list()
    page <- 1
    repeat {
      res <- tryCatch(fetch_notices_page(month_start, month_end, page),
                      error = function(e) NULL)
      if (is.null(res)) break
      rows <- res$notices %||% res$results %||% res$content
      if (is.null(rows) || NROW(rows) == 0) break
      pages[[page]] <- rows
      if (NROW(rows) < 100) break   # last page
      page <- page + 1
    }
    bind_rows(pages)
  })
}

`%||%` <- function(a, b) if (is.null(a)) b else a
```

### Filings over time

```{r notices}
months <- seq(as.Date(paste0(params$start_year, "-01-01")),
              floor_date(Sys.Date(), "month") - months(1),
              by = "month")

notices_raw <- map(months, fetch_month) |> bind_rows()

stopifnot(NROW(notices_raw) > 0)

# One company can file more than once — an initial filing and later updates
# are both ESR entries. They are counted separately, because the update rate
# is itself informative: a register that is filled once and never revised is
# not current, whatever its coverage looks like.
ubo_notices <- notices_raw |>
  filter(map_lgl(registrationNoticeTypes,
                 ~ any(str_detect(as.character(.x), params$notice_code)))) |>
  transmute(
    business_id = businessId,
    reg_date    = as.Date(registrationDate),
    company_form = companyForm
  ) |>
  arrange(business_id, reg_date) |>
  group_by(business_id) |>
  mutate(filing_seq = row_number()) |>
  ungroup()

message("UBO filings: ", nrow(ubo_notices),
        " across ", n_distinct(ubo_notices$business_id), " companies")
```

```{r timeline}
#| fig-height: 4
monthly <- ubo_notices |>
  mutate(month = floor_date(reg_date, "month"),
         kind = if_else(filing_seq == 1, "First filing", "Update")) |>
  count(month, kind)

ggplot(monthly, aes(month, n, fill = kind)) +
  geom_col() +
  geom_vline(data = filter(policy_events, country == "FI"),
             aes(xintercept = date), linetype = "dashed") +
  labs(x = NULL, y = "Filings per month", fill = NULL,
       caption = "Dashed lines: statutory dates identified in Part 1") +
  theme_minimal()
```

### Coverage: the denominator

```{r denominator}
# Counting filings alone says nothing. The question is what share of the
# companies subject to the obligation have met it, which needs the full
# register as a denominator. PRH publishes the entire set of active
# companies as a single download.
companies <- cached("companies", {
  # TODO: confirm the bulk download URL from the YTJ API page. The bulk file
  # is the intended route; paging the whole register through the API is both
  # slower and ruder.
  stop("Set the YTJ bulk download URL before running this chunk.")
})

obliged_forms <- c("OY", "OYJ", "OSK")  # limited companies and co-operatives

obliged <- companies |>
  filter(company_form %in% obliged_forms)

coverage <- obliged |>
  mutate(has_filed = business_id %in% ubo_notices$business_id) |>
  summarise(
    companies = n(),
    filed     = sum(has_filed),
    share     = filed / companies
  )

coverage |>
  mutate(share = scales::percent(share, accuracy = 0.1)) |>
  kable(caption = "Share of obliged companies with a registered UBO filing")
```

### Who has not filed

```{r profile}
# The profile of non-filing is the part a compliance team can act on. None
# of these variables identifies a person; they are register attributes of
# the legal entity.
profile <- obliged |>
  mutate(
    has_filed = business_id %in% ubo_notices$business_id,
    age_years = as.numeric(Sys.Date() - as.Date(registration_date)) / 365.25,
    age_band  = cut(age_years, c(0, 2, 5, 10, 20, Inf),
                    labels = c("<2", "2-5", "5-10", "10-20", "20+"))
  )

profile |>
  group_by(company_form, age_band) |>
  summarise(n = n(), not_filed = mean(!has_filed), .groups = "drop") |>
  filter(n >= 50) |>
  mutate(not_filed = scales::percent(not_filed, accuracy = 0.1)) |>
  kable(caption = "Non-filing rate by company form and age")
```

### Did the sanctions work?

```{r response}
# The test: filing volume should jump at the dates Part 1 identified and
# not at arbitrary ones. A simple before/after comparison around each event,
# against the surrounding months as a baseline, is enough to see whether the
# response is real. Anything more elaborate would be precision the data does
# not support.
window_effect <- function(event_date, window = 3) {
  before <- ubo_notices |>
    filter(reg_date >= event_date %m-% months(window), reg_date < event_date) |>
    nrow()
  after <- ubo_notices |>
    filter(reg_date >= event_date, reg_date < event_date %m+% months(window)) |>
    nrow()
  tibble(event_date, before, after, ratio = after / pmax(before, 1))
}

filter(policy_events, country == "FI") |>
  pull(date) |>
  map_dfr(window_effect) |>
  mutate(ratio = round(ratio, 2)) |>
  kable(caption = "Filings in the three months before and after each statutory date")
```

## What this means for an obliged entity

*To be written once the numbers exist. The argument to make: the share of
the register that has never filed is a counterparty risk signal available
to anyone in Finland and to nobody in Sweden, Denmark or Norway, and it is
derived from metadata rather than from the protected data itself.*

## Limitations

- A registered filing is not a correct filing. This measures whether a
  company met a formal obligation, not whether the information it supplied
  is accurate, and the register's own supervisory markings for deficient
  data are not public.
- Companies struck off the register for failing to file leave the
  denominator, which flatters the coverage figure over time. The direction
  of that bias is known; its size is not.
- The obligation does not apply uniformly to every company form, so the
  denominator depends on a legal judgement made in Part 1 rather than on
  the data.
- Nothing here identifies a beneficial owner, and nothing here could. That
  is the point: the measurement is built from the existence of filings, not
  their content.
